### 2021-05-07 article.Rnw
add old vignette

<<echo=FALSE, results='hide'>>=
# This should be set to FALSE in case you want to redo the sampling as well.
# Default is TRUE to save CRAN some time.
usePreCalcResults <- FALSE

# Default here is FALSE because the object created is rather large
usePreCalcResultsExtra <- FALSE
@

%\VignetteIndexEntry{Dealing with Stochastic Volatility in Time Series Using the R Package stochvol}

%\VignetteKeyword{Bayesian inference}
%\VignetteKeyword{Markov chain Monte Carlo (MCMC)}
%\VignetteKeyword{heteroskedasticity}
%\VignetteKeyword{SV}
%\VignetteKeyword{GARCH}
%\VignetteKeyword{forecasting}
%\VignetteKeyword{predictive density}
%\VignetteKeyword{predictive Bayes factor}
%\VignetteKeyword{exchange rates}

\documentclass[article, nojss]{jss}
%\documentclass[article]{jss}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{textcomp}
\usepackage{tablefootnote}

\input{macros.tex}

\graphicspath{{extrafig/}}

\interfootnotelinepenalty=10000

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% almost as usual
\author{Gregor Kastner\\WU Vienna University of Economics and Business}
\title{Dealing with Stochastic Volatility in Time Series Using the \proglang{R} Package \pkg{stochvol}}
\Plaintitle{Dealing with Stochastic Volatility in Time Series Using the R Package stochvol} %% without formatting

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Gregor Kastner} %% comma-separated

%% an abstract and keywords
\Abstract{
The \proglang{R} package \pkg{stochvol} provides a fully Bayesian implementation of heteroskedasticity modeling within the framework of stochastic volatility.
It utilizes Markov chain Monte Carlo (MCMC) samplers to conduct inference by obtaining draws from the posterior distribution of parameters and latent variables which can then be used for predicting future volatilities. The package can straightforwardly be employed as a stand-alone tool; moreover, it allows for easy incorporation into other MCMC samplers. The main focus of this paper is to show the functionality of \pkg{stochvol}. In addition, it provides a brief mathematical description of the model, an overview of the sampling schemes used, and several illustrative examples using exchange rate data.
}
\Keywords{Bayesian inference, Markov chain Monte Carlo (MCMC), heteroskedasticity, SV, GARCH, forecasting, predictive density, predictive Bayes factor, exchange rates}
%% at least one keyword must be supplied

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Gregor Kastner\\
  Institute for Statistics and Mathematics\\
  Department of Finance, Accounting and Statistics\\
  WU Vienna University of Economics and Business\\
  Welthandelsplatz 1, Building D4, Level 4\\
  1020 Vienna, Austria\\
  Telephone: +43/1/31336-5593\\
  Fax: +43/1/31336-90-5593\\
  E-mail: \email{gregor.kastner@wu.ac.at}\\
  URL: \url{http://statmath.wu.ac.at/~kastner/}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

<<setup, include=FALSE, cache=FALSE>>=
knitr::render_sweave()
knitr::opts_chunk$set(prompt = TRUE,
                      fig.show = "hide",
                      fig.keep = "high",
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE,
                      echo = FALSE,
                      cache = FALSE,
                      fig.path = "Figures/jss2014-",
                      tidy = FALSE)
base::options(continue = "+  ", prompt = "R> ", width = 70,
  useFancyQuotes = FALSE)
plotevery <- 10L # can be used for trace plots to make vignette smaller
@

\shortcites{r:mvt}
%% include your article here, just as usual
%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.

\section*{Preface}
This vignette corresponds to an article of the same name which is published in the Journal of Statistical Software. The code base of the vignette has been updated to reflect the changes in recent versions of \stochvol{}. To cite, please use \cite{kas:dea}. Further information about citing \pkg{stochvol} can be obtained in \proglang{R} by installing the package, e.g., through \code{install.packages("stochvol")}, and calling \code{citation("stochvol")}.

\section{Introduction}
Returns -- in particular financial returns -- are commonly analyzed by estimating and predicting potentially time-varying volatilities. This focus has a long history, dating back at least to \cite{mar:por} who investigates portfolio construction with optimal expected return-variance trade-off. In his article, he proposes rolling-window-type estimates for the instantaneous volatilities, but already then recognizes the potential for ``better methods, which take into account more information''.

One approach is to model the evolution of volatility deterministically, i.e., through the (G)ARCH class of models. After the groundbreaking papers of \cite{eng:aut} and \cite{bol:gen}, these models have been generalized in numerous ways and applied to a vast amount of real-world problems. As an alternative, \cite{tay:fin} proposes in his seminal work to model the volatility probabilistically, i.e., through a state-space model where the logarithm of the squared volatilities -- the latent states -- follow an autoregressive process of order one. Over time, this specification became known as the \emph{stochastic volatility (SV)} model. Even though several papers \citep[e.g.,][]{jac-etal:bayJBES, ghy-etal:sto, kim-etal:sto} provide early evidence in favor of using SV, these models have found comparably little use in applied work. This obvious discrepancy is discussed in \cite{bos:rel} who points out two reasons: the variety (and potential incompatibility) of estimation methods for SV models -- whereas the many variants of the GARCH model have basically a single estimation method -- and the lack of standard software packages implementing these methods.

%While there are literally thousands of applications of GARCH, for SV, this number is far lower. Two reasons for this relative lack of applied work using SV are apparent. First, there are as of yet no standard packages for estimating SV models, whereas for GARCH, most statistical packages have a wealth of options for incorporating GARCH effects. A second difference seems to be that GARCH has many variants of the model \citep{bol:glo}, with basically a single estimation method for all of them. For SV, there are few variants of the model, but a full series of estimation methods.

 In \cite{kas-fru:anc}, the former issue is thoroughly investigated and an efficient MCMC estimation scheme
%using techniques such as sampling AWOL \citep[``all without a loop'', adapted from][]{rue:fas, mcc-etal:sim} and ASIS \citep[``ancillarity-sufficiency interweaving strategy'', adapted from][]{yu-men:cen}, an
is proposed. The paper at hand and the corresponding package \pkg{stochvol} \citep{r:sto} for \proglang{R} \citep{r:r} are crafted to cope with the latter problem: the apparent lack of ready-to-use software packages for efficiently estimating SV models.

\section{Model specification and estimation}

We begin by briefly introducing the model and specifying the notation used in the remainder of the paper. Furthermore, an overview of Bayesian parameter estimation via Markov chain Monte Carlo (MCMC) methods is given.

\subsection{The SV model}

Let $\yb = (y_1, y_2, \dots, y_n)^\top$ be a vector of returns with mean zero. The intrinsic feature of the SV model is that each observation $y_t$ is assumed to have its ``own'' contemporaneous variance $\e^{\hsv_t}$, thus relaxing the usual assumption of homoskedasticity. In order to make the estimation of such a model feasible,
%(note that there are just as many data points as there are variances!)
this variance is not allowed to vary unrestrictedly with time. Rather, its logarithm is assumed to follow an autoregressive process of order one. Note that this feature is fundamentally different to GARCH-type models where the time-varying volatility is assumed to follow a deterministic instead of a stochastic evolution.

The SV model can thus be conveniently expressed in hierarchical form. In its centered parameterization, it is given through
\begin{eqnarray}
 \label{c1}
 y_{t}|\hsv_t &\sim&  \Normal{0, \exp{\hsv_{t}}},%\error_{t},% \qquad \error_{t} \sim \Normal{0,1},
 \\
 \label{c2}
 \hsv_{t}|\hsv_{t-1},\mupar, \phipar, \sigmapar&\sim& \Normal{\mupar +  \phipar (\hsv_{t-1}- \mupar ),  \sigmapartwo},\\% \etat_{t},% \qquad  \etat_{t} \sim \Normal{0,1},
 \label{c3}
\hsv_0|\mu,\phi,\sigmapar &\sim& \Normal{\mu,\sigmapartwo/(1-\phi^2)},
\end{eqnarray}

where $\Normal{\mu, \sigmapartwo}$ denotes the normal distribution with mean $\mu$ and variance $\sigmapartwo$. We refer to $\allpara = (\mupar, \phipar, \sigmapar)^\top$ as the vector of \emph{parameters}: the \emph{level} of log-variance $\mupar$, the \emph{persistence} of log-variance $\phipar$, and the \emph{volatility} of log-variance $\sigmapar$. The process $\hsvm=(\hsv_{0}, \hsv_{1}, \ldots,\hsv_{n})$ appearing in Equation~\ref{c2} and Equation~\ref{c3} is unobserved and usually interpreted as the latent time-varying \emph{volatility process} (more precisely, the log-variance process). Note that the initial state $\hsv_0$ appearing in Equation~\ref{c3} is distributed according to the stationary distribution of the autoregressive process of order one.

\subsection{Prior distribution}
\label{priors}

To complete the model setup, a prior distribution for the parameter vector $\allpara$ needs to be specified. Following \cite{kim-etal:sto}, we choose independent components for each parameter, i.e., $p(\allpara) = p(\mupar)p(\phipar)p(\sigmapar)$.

The level $\mupar \in \mathbb{R}$ is equipped with the usual normal prior $\mupar \sim \Normal{b_{\mupar}, B_{\mupar}}$. In practical applications, this prior is usually chosen to be rather uninformative, e.g., through setting $b_\mupar = 0$ and $B_\mupar \geq 100$ for daily log returns. Our experience with empirical data is that the exact choice is usually not very influential; see also Section~\ref{priorconfig}.

For the persistence parameter $\phipar \in (-1,1)$, we choose $(\phi+1)/2 \sim \Betadis{a_0, b_0}$, implying
\begin{eqnarray}
 \label{beta_transformed}
 p(\phi) = \frac{1}{2\Betafun{a_0,b_0}}\left ( \frac{1+\phi}{2}\right ) ^{a_0-1}\left ( \frac{1-\phi}{2}\right ) ^{b_0-1},
\end{eqnarray}
where $a_0$ and $b_0$ are positive hyperparameters and $\Betafun{x,y} = \int_0^1t^{x-1}(1-t)^{y-1}\,dt$ denotes the beta function.
Clearly, the support of this distribution is the interval $(-1, 1)$; thus, stationarity of the autoregressive volatility process is guaranteed. Its expected value and variance are given through the expressions
\begin{eqnarray*}
 E(\phipar) &=& \frac{2a_0}{a_0+b_0}-1,\\
 V(\phipar) &=& \frac{4a_0b_0}{(a_0+b_0)^2(a_0+b_0+1)}.
\end{eqnarray*}
This obviously implies that the prior expectation of $\phipar$ depends only on the ratio $a_0:b_0$. It is greater than zero if and only if $a_0 > b_0$ and smaller than zero if and only if $a_0 < b_0$. For a fixed ratio $a_0:b_0$, the prior variance decreases with larger values of $a_0$ and $b_0$. The uniform distribution on $(-1,1)$ arises as a special case when $a_0=b_0=1$. For financial datasets with not too many observations (i.e., $n \lesssim 1000$), the choice of the hyperparameters $a_0$ and $b_0$ can be quite influential on the shape of the posterior distribution of $\phipar$. In fact, when the underlying data-generating process is (near-)homoskedastic, the volatility of log-variance $\sigmapar$ is (very close to) zero and thus the likelihood contains little to no information about $\phipar$. Consequently, the posterior distribution of $\phipar$ is (almost) equal to its prior, no matter how many data points are observed. For some discussion about this issue, see, e.g., \cite{kim-etal:sto} who choose $a_0=20$ and $b_0=1.5$, implying a prior mean of $0.86$ with a prior standard deviation of $0.11$ and thus very little mass for nonpositive values of $\phi$.

For the volatility of log-variance $\sigmapar \in \mathbb{R}^+$, we choose $\sigmapartwo \sim B_{\sigmapar}\times \chi^2_1=\Gammad{1/2,1/2B_{\sigmapar}}$. This choice is motivated by \cite{fru-wag:sto} who equivalently stipulate the prior for $\pm \sqrt { \sigmapartwo}$ to follow a centered normal distribution, i.e., $\pm \sqrt { \sigmapartwo} \sim \Normal{0,B_{\sigmapar}}$. As opposed to the more common Inverse-Gamma prior for $\sigmapartwo$, this prior is not conjugate in the usual sampling scheme. However, it does not bound $\sigmapartwo$ away from zero a priori. The choice of the hyperparameter $B_{\sigmapar}$ turns out to be of minor influence in empirical applications as long as it is not set too small.

\subsection{MCMC sampling}

An MCMC algorithm such as the one implemented in the package \pkg{stochvol} provides its user with draws from the posterior distribution of the desired random variables; in our case with the latent log-variances $\hsvm$ and the parameter vector $\allpara$. Because these draws are usually dependent, Bayesian inference via MCMC may require careful design of the algorithm and attentive investigation of the draws obtained.

One key feature of the algorithm used in this package is the joint sampling of all instantaneous volatilities ``all without a loop'' (AWOL), a technique going back at least to \cite{rue:fas} and discussed in more detail in \cite{mcc-etal:sim}. Doing so reduces correlation of the draws significantly and requires auxiliary finite mixture approximation of the errors as in \cite{kim-etal:sto} or \cite{omo-etal:sto}.

In order to avoid the cost of code interpretation within each MCMC iteration, the core computations are implemented in \proglang{C}. Their output is interfaced to \proglang{R} via the \pkg{Rcpp} package \citep{edd-fra:rcp}; there, the convenience functions and the user interface are implemented. This combination allows to make use of the well-established and widely accepted ease-of-use of \proglang{R} and its underlying functional programming paradigm. Moreover, existing frameworks for analyzing MCMC output such as \pkg{coda} \citep{plu-etal:cod} as well as high-level visualization tools can easily be used. Last but not least, users with a basic knowledge of \proglang{R} can use the package with a very low entry cost. Nevertheless, despite all these convenience features, the package profits from a highly optimized machine code generated by a compiler at package build time, thus providing acceptable runtime even for larger datasets.

A novel and crucial feature of the algorithm implemented in \pkg{stochvol} is the usage of a variant of the ``ancillarity-sufficiency interweaving strategy'' (ASIS) which has been brought forward in the general context of state-space models by \cite{yu-men:cen}. ASIS exploits the fact that for certain parameter constellations sampling efficiency improves substantially when considering a non-centered version of a state-space model.
This is commonly referred to as a reparameterization issue and dealt with extensively in scholarly literature; for an early reference see, e.g., \cite{hil-smi:par}.
For the model at hand, a move of this kind can be achieved by transferring the level of log-variance $\mupar$ and/or the volatility of log-variance $\sigmapar$ from the state process (Equations~\ref{c2}~and~\ref{c3}) to the observation process (Equation~\ref{c1}) through a simple reparameterization of $\hsvm$. However, in the case of the SV model, it turns out that no single superior parameterization exists. Rather, for some underlying processes, the standard parameterization yields superior results, while for other processes non-centered versions are better. To overcome this issue, the parameter vector $\allpara$ is sampled twice: once in the centered and once in a noncentered parameterization. This method of ``combining best of different worlds'' allows for efficient inference regardless of the underlying process with one algorithm. For more details about the algorithm and empirical results concerning sampling efficiency, see \cite{kas-fru:anc}.

\section[The stochvol package]{The \pkg{stochvol} package}

The usual stand-alone approach to fitting SV models with \pkg{stochvol} exhibits the following workflow: (1) Prepare the data, (2) specify the prior distributions and configuration parameters, (3) run the sampler, (4) assess the output and display the results. All these steps are described in more detail below, along with a worked example. For a stepwise incorporation of SV effects into other MCMC samplers, see Section~\ref{other}.

\subsection{Preparing the data}
\label{prep}

The core sampling function \code{svsample} expects its input data \code{y} to be a numeric vector of returns without any missing values (\code{NA}s) and throws an error if provided with anything else. In the case that \code{y} contains zeros, a warning is issued and a small offset constant of size \code{sd(y)/10000} is added to the squared returns before doing the auxiliary mixture sampling \citep[cf.][]{omo-etal:sto}. However, we generally recommend to avoid zero returns altogether, e.g., by demeaning them beforehand.

Below is an illustration of how to prepare data by using the sample dataset \code{exrates}\footnote{The dataset -- obtained from the European Central Bank's Statistical Data Warehouse -- contains the daily bilateral prices of one euro in 23 currencies from January 3, 2000 until April 4, 2012. Conversions to New Turkish lira and Fourth Romanian leu have been incorporated. See \code{?exrates} for more information.} included in the package. Figure~\ref{fig1} provides a visualization of a time series from this dataset.

<<usd1, echo=TRUE, fig.width=11, fig.height=6>>=
set.seed(123)
library("stochvol")
data("exrates")
ret <- logret(exrates$USD, demean = TRUE)
par(mfrow = c(2, 1), mar = c(1.9, 1.9, 1.9, 0.5), mgp = c(2, 0.6, 0))
plot(exrates$date, exrates$USD, type = "l",
  main = "Price of 1 EUR in USD")
plot(exrates$date[-1], ret, type = "l", main = "Demeaned log returns")
@
\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=\textwidth]{Figures/jss2014-usd1}
 \caption{Visualization of EUR-USD exchange rates included in the \pkg{stochvol} package.}
 \label{fig1}
\end{center}
\end{figure}

Additionally to real-world data, \pkg{stochvol} also has a built-in data generator \code{svsim}. This function simply produces realizations of an SV process and returns an object of class \code{svsim} which has its own \code{print}, \code{summary}, and \code{plot} methods. Exemplary code using \code{svsim} is given below and the particular instance of this simulated series is displayed in Figure~\ref{fig2}.

<<usd2, echo=TRUE, fig.width=11, fig.height=6>>=
sim <- svsim(500, mu = -9, phi = 0.99, sigma = 0.1)
par(mfrow = c(2, 1))
plot(sim)
@
\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=\textwidth]{Figures/jss2014-usd2}
 \caption{Visualization of a simulated time series as provided by the default \code{plot} method.}
 \label{fig2}
\end{center}
\end{figure}

 \subsection{Specifying prior distributions and configuration parameters}
 \label{priorconfig}

 After preparing the data vector \code{y}, the user needs to specify the prior hyperparameters for the parameter vector $\allpara = (\mupar, \phipar, \sigmapar)^\top$ -- see also Section~\ref{priors} -- and some configuration parameters. The appropriate values are passed to the main sampling function \code{svsample} as arguments which are described below.

 The argument \code{priormu} is a vector of length $2$, containing mean and standard deviation of the normal prior for the level of the log-variance $\mupar$. A common strategy is to choose a vague prior here, e.g., \code{c(0, 100)}, because the likelihood usually carries enough information about this parameter. If one prefers to use (slightly) informative priors, e.g., to avoid outlier draws of $\mupar$, one must pay attention to whether \emph{log returns} or \emph{percentage log returns} are analyzed. Assuming daily data, log returns commonly have an unconditional variance of $0.0001$ or less and thus the level on the log scale $\mupar$ lies around $\log(0.0001) \approx -9$. Percentage log returns, on the other hand, have the $100^2$-fold unconditional variance (around $1$) which implies a level of $\log(1) = 0$. Choices in the literature include \code{c(0, 10)} \citep{jac-etal:bayJE}, \code{c(0, 5)} \citep{yu:on}, \code{c(0, sqrt(10))} \citep{kim-etal:sto, mey-yu:bug} or \code{c(0, 1)} \citep{omo-etal:sto}. Note that most of these choices are quite informative and clearly designed for percentage log returns.

 For specifying the prior hyperparameters for the persistence of log-variance, $\phipar$, the argument \code{priorphi} may be used. It is again a vector of length $2$, containing $a_0$ and $b_0$ specified in Equation~\ref{beta_transformed}. As elaborated in Section~\ref{priors}, these values can possibly be quite influential, thus we advise to choose them carefully and study the effects of different choices. The default is currently given through \code{c(5, 1.5)}, implying a prior mean of $0.54$ and a prior standard deviation of $0.31$.

 The prior variance of log-variance hyperparameter $B_\sigmapar$ may be controlled through \code{priorsigma}. This argument defaults to \code{1} if not provided by the user. As discussed in Section~\ref{priors}, the exact choice of this value is usually not very influential in typical applications. In general, it should not be set too small unless there is a very good reason, e.g., explicit prior knowledge, to do so.

For specifying the size of the burn-in, the parameter \code{burnin} is provided. It is the amount of MCMC iterations that are run but discarded to ensure convergence to the stationary distribution of the chain. The current default value for this parameter is $1000$ which has turned out to suffice in most situations. Nevertheless, the user is encouraged to check convergence carefully; see Section~\ref{check} for more details. The amount of iterations which are run after burn-in can be specified through the parameter \code{draws}, currently defaulting to $10\,000$. Consequently, the sampler is run for a total of \code{burnin + draws} iterations.

 Three thinning parameters are available which all are $1$ if not specified otherwise. The first one, \code{thinpara}, is the denominator in the fraction of parameter draws (i.e., draws of $\allpara$) that are stored. E.g., if \code{thinpara} equals $10$, every $10$th draw is kept. The default parameter thinning value of $1$ means that all draws are saved. The second thinning parameter, \code{thinlatent}, acts in the same way for the latent variables $\hsvm$. The third thinning parameter, \code{thintime}, refers to thinning with respect to the time dimension of the latent volatility.
In the case that \code{thintime} is greater than 1, not all elements of $\hsvm$ are stored, e.g., for \code{thintime} equaling $10$, only the draws of $\hsv_1, \hsv_{11}, \hsv_{21}, \dots$ (and $\hsv_0$) are kept.

Another configuration argument is \code{quiet} which defaults to \code{FALSE}. If set to \code{TRUE}, all output during sampling (progress bar, status messages) is omitted. The arguments \code{startpara} and \code{startlatent} are optional starting values for the parameter vector $\allpara$ and the latent variables $\hsvm$, respectively. All other configuration parameters are summarized in the argument \code{expert}, because it is not very likely that the end-user needs to mess with the defaults.\footnote{Examples of configurations that can be changed with the \code{expert}-argument include the specification of the (baseline) parameterization (either \emph{centered} or \emph{noncentered}) and the possibility to turn off interweaving. Moreover, some algorithmic details such as the number of blocks used for the parameter updates or the possibility of using a random walk Metropolis-Hastings proposal (instead of the default independence proposal) can be found here.} Please refer to the package documentation and \cite{kas-fru:anc} for details.

Any further arguments (\code{\dots}) are forwarded to \code{updatesummary}, controlling the type of summary statistics that are calculated for the posterior draws.

\subsection{Running the sampler}

At the heart of the package \pkg{stochvol} lies the function \code{svsample} which serves as an \proglang{R}-wrapper for the actual sampler coded in \proglang{C}. Exemplary usage of this function is given in the code snipped below, along with the default output.

<<echo=TRUE>>=
res <- svsample(ret, priormu = c(-10, 1), priorphi = c(20, 1.1),
  priorsigma = 0.1)
@

<<results='hide'>>=
if (usePreCalcResultsExtra) {
 load("vignette_sampling_draws1.RData")
} else {
 res <- svsample(ret, priormu = c(-10, 1), priorphi = c(20, 1.1), priorsigma = .1)
 #save(res, file = "vignette_sampling_draws1.RData")
}
@

\begin{Soutput}
Calling GIS_C MCMC sampler with 11000 iter. Series length is 3139.

  0% [+++++++++++++++++++++++++++++++++++++++++++++++++++] 100%

Timing (elapsed): 12.92 seconds.
851 iterations per second.

Converting results to coda objects... Done!
Summarizing posterior draws... Done!
\end{Soutput}

As can be seen, this function calls the main MCMC sampler and converts its output to \pkg{coda}-compatible objects. The latter is done mainly for reasons of compatibility and in order to have straightforward access to the convergence diagnostics checks implemented there. Moreover, some summary statistics for the posterior draws are calculated. The return value of \code{svsample} is an object of type \code{svdraws} which is a named list with eight elements, holding (1) the parameter draws in \code{para}, (2) the latent log-volatilities in \code{latent}, (3) the initial latent log-volatility draw in \code{latent0}, (4) the data provided in \code{y}, (5) the sampling runtime in \code{runtime}, (6) the prior hyperparameters in \code{priors}, (7) the thinning values in \code{thinning}, and (8) summary statistics of these draws, alongside some common transformations thereof, in \code{summary}.

 \subsection{Assessing the output and displaying the results}
 \label{check}

 Following common practice, \code{print} and \code{summary} methods are available for \code{svdraws} objects. Each of these has two optional parameters, \code{showpara} and \code{showlatent}, specifying which output should be displayed. If \code{showpara} is \code{TRUE} (the default), values/summaries of the parameter draws are shown. If \code{showlatent} is \code{TRUE} (the default), values/summaries of the latent variable draws are shown. In the example below, the summary for the parameter draws only is displayed.

<<echo=TRUE>>=
summary(res, showlatent = FALSE)
@

There are several plotting functions specifically designed for objects of class \code{svsample} which are described in the following paragraphs.
 \begin{enumerate}
  \item[(1)] \code{volplot}: Plots posterior quantiles of the latent volatilities in percent, i.e., empirical quantiles of the posterior distribution of $100\exp(h_t/2)$, over time. Apart from the mandatory \code{svsample}-object itself, this function takes several optional arguments. Only some are mentioned here; for an exhaustive list please see the corresponding help document accessible through \code{?volplot} or \code{help(volplot)}. Selected optional arguments that are commonly used include \code{forecast} for $n$-step-ahead volatility prediction, \code{dates} for labels on the $x$-axis, alongside some graphical parameters. The code snipped below shows a typical example and Figure~\ref{fig3} displays its output.

<<paropar, include=FALSE>>=
par(mfrow = c(1, 1))
@

<<usd3, echo=TRUE, fig.width=11, fig.height=7>>=
volplot(res, forecast = 100, dates = exrates$date[-1])
@

\begin{figure}[!htp]
\begin{center}
 \includegraphics[width=\textwidth]{Figures/jss2014-usd3}
 \caption{Visualization of estimated contemporaneous volatilities of EUR-USD exchange rates, as provided by \code{volplot}. If not specified otherwise, posterior medians and 5\%/95\% quantiles are plotted. The dotted lines on the right indicate predicted future volatilities.}
 \label{fig3}
\end{center}
\end{figure}

In case the user wants to display different posterior quantiles, the \code{updatesummary} function has to be called first. See the code below for an example and Figure~\ref{fig4} for the corresponding plot.

<<usd4, echo=TRUE, fig.width=11, fig.height=7>>=
res <- updatesummary(res, quantiles = c(0.01, 0.1, 0.5, 0.9, 0.99))
volplot(res, forecast = 100, dates = exrates$date[-1])
@

\begin{figure}[!htp]
\begin{center}
 \includegraphics[width=\textwidth]{Figures/jss2014-usd4}
 \caption{As above, now with medians (black line) and 1\%/10\%/90\%/99\% quantiles (gray lines). This behavior can be achieved through a preceding call of \code{updatesummary}.}
 \label{fig4}
\end{center}
\end{figure}

 \item[(2)] \code{paratraceplot}: Displays trace plots for the parameters contained in $\allpara$. Note that the burn-in has already been discarded. Figure~\ref{fig5} shows an example.

<<usd5, echo=TRUE, fig.width=8, fig.height=5>>=
par(mfrow = c(3, 1))
paratraceplot(res)
@

\begin{figure}[!htp]
\begin{center}
 \includegraphics[width=\textwidth]{Figures/jss2014-usd5}
 \caption{Trace plots of posterior draws for the parameters $\mupar, \phipar, \sigmapar$.}
 %The underlying dataset is again EUR-USD exchange rates.}
 \label{fig5}
\end{center}
 \end{figure}

\item[(3)] \code{paradensplot}: Displays a kernel density estimate for the parameters contained in $\allpara$. If the argument \code{showobs} is \code{TRUE} (which is the default), individual posterior draws are indicated through a rug, i.e., short vertical lines along the $x$-axis. For quicker drawing of large posterior samples, this argument should be set to \code{FALSE}. If the argument \code{showprior} is \code{TRUE} (which is the default), the prior distribution is indicated through a dashed gray line. Figure~\ref{fig6} shows a sample output for the EUR-USD exchange rates obtained from the \code{exrates} dataset.

<<usd6, echo=TRUE, fig.width=8, fig.height=3.5>>=
par(mfrow = c(1, 3))
paradensplot(res, showobs = FALSE)
@

\begin{figure}[!htp]
\begin{center}
 \includegraphics[width=\textwidth]{Figures/jss2014-usd6}
 \caption{Posterior density estimates (black solid lines) along with prior densities (dashed gray lines). Individual posterior draws are indicated by the underlying rug.}
 \label{fig6}
\end{center}
\end{figure}
\end{enumerate}

The generic \code{plot} method for \code{svdraws} objects combines all above plots into one plot. All arguments described above can be used. See \code{?plot.svsample} for an exhaustive summary of possible arguments and Figure~\ref{fig7} for an example.

<<usd7, echo=TRUE, fig.width=9, fig.height=7>>=
plot(res, showobs = FALSE)
@

\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=\textwidth]{Figures/jss2014-usd7}
 \caption{Illustration of the default \code{plot} method for \code{svdraws}-objects. This visualization combines \code{volplot} (Figure~\ref{fig4}), \code{traceplot} (Figure~\ref{fig5}), and \code{paradensplot} (Figure~\ref{fig6}) into one single plot.}
 \label{fig7}
\end{center}
\end{figure}

For extracting standardized residuals, the \code{residuals}/\code{resid} method can be used on a given \code{svdraws} object. With the optional argument \code{type}, the type of summary statistic may be specified. Currently, \code{type} is allowed to be either \code{"mean"} or \code{"median"}, where the former corresponds to the default value. This method returns a real vector of class \code{svresid} which contains the requested summary statistic of standardized residuals for each point in time. There is also a \code{plot} method available, providing the option of comparing the standardized residuals to the original data when given through the argument \code{origdata}. See the code below for an example and Figure~\ref{fig8} for the corresponding output.

<<usd8, echo=TRUE, fig.width=9, fig.height=7>>=
myresid <- resid(res)
plot(myresid, ret)
@

\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=\textwidth]{Figures/jss2014-usd8}
 \caption{Mean standardized residual plots for assessing the model fit, as provided by the corresponding \code{plot} method. The dashed lines in the bottom left panel indicate the 2.5\%/97.5\% quantiles of the standard normal distribution.}
 \label{fig8}
\end{center}
\end{figure}

\section[Using stochvol within other samplers]{Using \pkg{stochvol} within other samplers}
\label{other}
We demonstrate how the \pkg{stochvol} package can be used to incorporate stochastic volatility into any given MCMC sampler. This is particularly easy when the sampler itself is coded in \proglang{R} or \proglang{C}/\proglang{C++} and applies, e.g., to many of the specialized procedures listed in the CRAN task view about Bayesian inference, available at \url{http://cran.r-project.org/web/views/Bayesian.html}. For the sake of simplicity, we explain the procedure using a ``hand-coded'' Gibbs-sampler for the Bayesian normal linear model with $n$ observations and $k=p-1$ predictors, given through
\begin{eqnarray}
 \label{reg}
\yb|\betab,\Sigmab \sim \Normal{\Xb\betab, \Sigmab}.
\end{eqnarray}
Here, $\yb$ denotes the $n\times1$ vector of responses, $\Xb$ is the $n \times p$ design matrix containing ones in the first column and the predictors in the others, and $\betab = (\beta_0, \beta_1, \dots, \beta_{p-1})^\top$ stands for the $p\times 1$ vector of regression coefficients. In the following sections, we discuss two specifications of the $n\times n$ error covariance matrix $\Sigmab$.

\subsection{The Bayesian normal linear model with homoskedastic errors}
\label{homo}
The arguably simplest specification of the error covariance matrix in Equation~\ref{reg} is given by $\Sigmab\equiv\sigmaepsextrasuper2\Ib$, where $\Ib$ denotes the $n$-dimensional unit matrix. This specification is used in many applications and commonly referred to as the linear regression model with homoskedastic errors. To keep things simple, let model parameters $\betab$ and $\sigmaepsextrasuper2$ be equipped with the usual conjugate prior $p(\betab, \sigmaepsextrasuper2) = p(\betab|\sigmaepsextrasuper2)p(\sigmaepsextrasuper2)$, where
\begin{eqnarray*}
\betab|\sigmaepsextrasuper2 &\sim& \Normal{\bb0, \sigmaepsextrasuper2\Bb0},\\
 \sigmaepsextrasuper2&\sim& \Gammainv{c_0, C_0}.
\end{eqnarray*}

Commonly, samples from the posterior distribution of this model are obtained through a Gibbs-algorithm, where draws are generated in turn from the full conditional distributions $\betab|\yb,\sigmaepsextrasuper2 \sim \Normal{\bpostb, \Bpostb}$ with
\[
\bpostb = \left(\XtX + \Bb0^{-1} \right)^{-1}\left( \Xb^\top\yb + \Bb0^{-1}\bb0 \right), \qquad
\Bpostb = \sigmaepsextrasuper2\left(\XtX + \Bb0^{-1} \right)^{-1},
\]

and $\sigmaepsextrasuper2|\yb,\betab \sim \Gammainv{c_n, C_n}$ with
\[
c_n = c_0 + \frac{n}{2} + \frac{p}{2}, \qquad
C_n = C_0 + \frac{1}{2}
 \left(
 (\yb - \Xb\betab)^\top(\yb - \Xb\betab) +
 (\betab - \bb0)^\top\Bb0^{-1}(\betab - \bb0)
 \right).
\]

In \proglang{R}, this can straightforwardly be coded as follows:
\begin{itemize}
 \item Set the seed to make results reproducible and simulate some data from the underlying model:
<<echo=TRUE>>=
set.seed(123456)
n <- 1000
beta.true <- c(0.1, 0.5)
sigma.true <- 0.01
X <- matrix(c(rep(1, n), rnorm(n, sd = sigma.true)), nrow = n)
y <- rnorm(n, X %*% beta.true, sigma.true)
@

\item Specify the size of the burn-in and the number of draws thereafter; set the prior parameters:
<<echo=TRUE>>=
burnin <- 100
draws <- 5000
b0 <- matrix(c(0, 0), nrow = ncol(X))
B0inv <- diag(c(10^-10, 10^-10))
c0 <- 0.001
C0 <- 0.001
@
 \item Pre-calculate some values outside the main MCMC loop:
<<echo=TRUE>>=
p <- ncol(X)
preCov <- solve(crossprod(X) + B0inv)
preMean <- preCov %*% (crossprod(X, y) + B0inv %*% b0)
preDf <- c0 + n/2 + p/2
@
\item Assign some storage space for holding the draws and set an initial value for $\sigmaepsextrasuper2$:
<<echo=TRUE>>=
draws1 <- matrix(NA_real_, nrow = draws, ncol = p + 1)
colnames(draws1) <- c(paste("beta", 0:(p-1), sep = "_"), "sigma")
sigma2draw <- 1
@
\item Run the main sampler: Iteratively draw from the conditional bivariate Gaussian distribution $\betab|\yb,\sigmaepsextrasuper2$, e.g., through the use of \pkg{mvtnorm} \citep{r:mvt}, and the conditional Inverse Gamma distribution $\sigmaepsextrasuper2|\yb,\betab$.
<<echo=TRUE>>=
for (i in -(burnin-1):draws) {
  betadraw <- as.numeric(mvtnorm::rmvnorm(1, preMean,
    sigma2draw * preCov))
  tmp <- C0 + 0.5 * (crossprod(y - X %*% betadraw) +
    crossprod((betadraw - b0), B0inv) %*% (betadraw - b0))
  sigma2draw <- 1 / rgamma(1, preDf, rate = tmp)
  if (i > 0) draws1[i, ] <- c(betadraw, sqrt(sigma2draw))
}
@
\item Calculate posterior means in order to obtain point estimates for the parameters:
<<echo=TRUE>>=
colMeans(draws1)
@

\item Visualize the draws through \pkg{coda}'s native \code{plot} method:

<<homo, echo=TRUE, fig.width=7.5, fig.height=6>>=
par(mar = c(3.1, 1.8, 1.9, .5), mgp = c(1.8, .6, 0))
plot(coda::mcmc(draws1[seq(1L, nrow(draws1), by = plotevery),]))
@

<<eval=FALSE, echo=TRUE>>=
plot(coda::mcmc(draws1))
@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\textwidth]{Figures/jss2014-homo}
\caption{Trace plots and kernel density estimates for some draws from the marginal posterior distributions in the regression model with heteroskedastic errors. Underlying data is simulated with $\betab^\text{true}=(0.1, 0.5)^\top$, $\sigmaepsextrasuper{\text{true}} = 0.01$, $n=1000$.}
\label{homodraws}
\end{center}
\end{figure}

\end{itemize}

\subsection{The Bayesian normal linear model with SV errors}
\label{hetero}

Instead of homoskedastic errors, we now specify the error covariance matrix in Equation~\ref{reg} to be $\Sigmab \equiv \diag{\e^{h_1}, \dots, \e^{h_n}}$, thus introducing nonlinear dependence between the observations due to the AR(1)-nature of $\hb$. Instead of cooking up an entire new sampler, we adapt the code from above utilizing the \pkg{stochvol} package.\footnote{Since \stochvol{} version 3.0, regression is a built-in feature of the function \code{svsample_fast_cpp} through its argument \code{designmatrix}.} To do so, we simply replace the sampling step of $\sigmaepsextrasuper{2}$ from an Inverse-Gamma distribution with a sampling step of $\allpara$ and $\hb$ through a call to \code{svsample_fast_cpp}.\footnote{In earlier version of the \pkg{stochvol} package, this function was called \code{svsample2}.} This function is a minimal-overhead version of the regular \code{svsample}. It provides the full sampling functionality of the original version but has slightly different default values, a simplified return value structure, and it does not perform costly input checks. Thus, it executes faster and is more suited for repeated calls. The drawback is that it needs to be used with proper care.\footnote{Erroneous or incompatible input values most likely result in run-time errors of compiled code, often implying a segmentation fault and the consecutive abnormal termination of \proglang{R}. This can render debugging tedious.} Note that the current draws of the variables need to be passed to the function through \code{startpara} and \code{startlatent}.

Here is how it goes:
\begin{itemize}
 \item Simulate some data:
<<echo=TRUE>>=
mu.true <- log(sigma.true^2)
phi.true <- 0.97
vv.true <- 0.3
simresid <- svsim(n, mu = mu.true, phi = phi.true, sigma = vv.true)
y <- X %*% beta.true + simresid$y
@
 \item Specify configuration parameters and prior values:
<<echo=TRUE>>=
draws <- 50000
burnin <- 1000
thinning <- 10
priors <- specify_priors(
  mu = sv_normal(-10, 2),
  phi = sv_beta(20, 1.5),
  sigma2 = sv_gamma(0.5, 0.5)
)
@
 \item Assign some storage space for holding the draws and set initial values:
<<echo=TRUE>>=
draws2 <- matrix(NA_real_, nrow = floor(draws / thinning),
  ncol = 3 + n + p)
colnames(draws2) <- c("mu", "phi", "sigma",
  paste("beta", 0:(p-1), sep = "_"), paste("h", 1:n, sep = "_"))
betadraw <- c(0, 0)
paradraw <- list(mu = -10, phi = 0.9, sigma = 0.2)
latentdraw <- rep(-10, n)
paranames <- names(paradraw)
@
 \item Run the main sampler, i.e., iteratively draw
  \begin{itemize}
   \item the latent volatilities/parameters by conditioning on the regression parameters and calling \code{svsample2},
   \item the regression parameters by conditioning on the latent volatilities and calling \code{rmvnorm}:
  \end{itemize}
<<eval=FALSE, echo=TRUE>>=
for (i in -(burnin-1):draws) {
  ytilde <- y - X %*% betadraw
  svdraw <- svsample_fast_cpp(ytilde, startpara = paradraw,
    startlatent = latentdraw, priorspec = priors)
  paradraw <- svdraw$para
  latentdraw <- drop(svdraw$latent)
  normalizer <- as.numeric(exp(-latentdraw / 2))
  Xnew <- X * normalizer
  ynew <- y * normalizer
  Sigma <- solve(crossprod(Xnew) + B0inv)
  mu <- Sigma %*% (crossprod(Xnew, ynew) + B0inv %*% b0)
  betadraw <- as.numeric(mvtnorm::rmvnorm(1, mu, Sigma))
  if (i > 0 && i %% thinning == 0) {
    draws2[i/thinning, 1:3] <- drop(paradraw)[paranames]
    draws2[i/thinning, 4:5] <- betadraw
    draws2[i/thinning, 6:(n+5)] <- latentdraw
  }
}
@

<<>>=
if (usePreCalcResults) {
 load("vignette_sampling_draws2.RData")
} else {
for (i in -(burnin-1):draws) {

 # draw latent volatilities and AR-parameters:
 ytilde <- y - X %*% betadraw
 svdraw <- svsample_fast_cpp(ytilde, startpara = paradraw,
   startlatent = latentdraw, priorspec = priors)
 paradraw <- svdraw$para
 latentdraw <- drop(svdraw$latent)

 # draw the betas:
 normalizer <- as.numeric(exp(-latentdraw/2))
 Xnew <- X * normalizer
 ynew <- y * normalizer
 Sigma <- solve(crossprod(Xnew) + B0inv)
 mu <- Sigma %*% (crossprod(Xnew, ynew) + B0inv %*% b0)
 betadraw <- as.numeric(mvtnorm::rmvnorm(1, mu, Sigma))

 # store the results:
 if (i > 0 & i %% thinning == 0) {
  draws2[i/thinning, 1:3] <- drop(paradraw)[paranames]
  draws2[i/thinning, 4:5] <- betadraw
  draws2[i/thinning, 6:(n+5)] <- latentdraw
 }
}
 draws2selection <- draws2[,4:8]
 #save(draws2selection, file = 'vignette_sampling_draws2.RData')
}
@
\item Finally, visualize and summarize (some) posterior draws:

<<eval=FALSE, echo=TRUE>>=
plot(coda::mcmc(draws2[, 4:7]))
@

<<hetero, fig.width=7.5, fig.height=8>>=
par(mar = c(3.1, 1.8, 1.9, .5), mgp = c(1.8, .6, 0))
plot(coda::mcmc(draws2selection[seq(1L, nrow(draws2selection), by = plotevery), 1:4]))
@

\begin{figure}[!ht]
 \begin{center}
\includegraphics[width=\textwidth]{Figures/jss2014-hetero}
\caption{Trace plots and kernel density estimates
%for draws from the marginal posterior distributions of the parameters
in the regression model with heteroskedastic errors. Data is simulated with $\betab^\text{true}=(0.1, 0.5)^\top$, $h_1^\text{true} = -8.28$, $h_2^\text{true} = -8.50$, $n=1000$.}
\label{heterodraws}
 \end{center}
\end{figure}

<<eval=FALSE, echo=TRUE>>=
colMeans(draws2[, 4:8])
@

<<echo=FALSE, echo=TRUE>>=
colMeans(draws2selection)
@

\end{itemize}

It should be noted that even though \code{svsample2} is considerably faster than \code{svsample}, the cost of interpreting this function in each MCMC iteration is still rather high (which lies in the nature of \proglang{R} as an interpreted language). Thus, as of package version \code{0.8-0}, a single-step \code{update} is also made available at the \proglang{C}/\proglang{C++} level for samplers coded there. For details, please consult the \code{NEWS} file in the package source; for an application case using this approach, see \cite{kas-etal:ana}.

\section{Illustrative predictive exercise}
In the following, we compare the performance of the Bayesian normal linear model with homoskedastic errors from Section~\ref{homo} with the Bayesian normal linear model with SV errors from Section~\ref{hetero} using the \code{exrates} dataset introduced in Section~\ref{prep}. As a benchmark, we also include the Bayesian normal linear model with GARCH(1,1) errors given through Equation~\ref{reg} with
\begin{eqnarray*}
 \label{garch1} \Sigmab &=& \diag{\sigma^2_1,\dots,\sigma^2_n},\\
 \label{garch2} \sigma^2_t &=& \alpha_0+\alpha_1\tilde y_{t-1}^2+\alpha_2\sigma_{t-1}^2,
\end{eqnarray*}
where time index $t=1,\dots,n$. In the second equation, $\tilde y_{t-1}$ denotes the past ``residual'', i.e., the $(t-1)$th element of $\tilde \yb = \yb - \Xb\betab$.

\subsection{Model setup}

We again use the daily price of 1 EUR in USD from January 3, 2000 to April 4, 2012, denoted by $\pb = (p_1, p_2, \dots, p_n)^\top$. This time however, instead of modeling log returns, we investigate the development of log levels by regression. Let $\yb$ contain the logarithm of all observations except the very first one, and let $\Xb$ denote the design matrix containing ones in the first column and lagged log prices in the second, i.e.,
\begin{eqnarray*}
\yb =
\begin{pmatrix}
 \log p_2\\
 \log p_3\\
 \vdots\\
 \log p_n
\end{pmatrix}, \quad
\Xb =
\begin{pmatrix}
1&\log p_1\\
1&\log p_2\\
\vdots&\vdots\\
1&\log p_{n-1}
\end{pmatrix}.
\label{regression}
\end{eqnarray*}
Note that this specification simply corresponds to an AR(1) model for the log prices with three different error specifications: homoskedastic, SV, GARCH(1,1). It is a generalization of directly dealing with demeaned log returns where $\beta_0$ is a priori assumed to be equal to the mean of the log returns and $\beta_1$ is held constant at $1$.
%
<<scatter, echo=TRUE, fig.width=10.2>>=
x <- log(exrates$USD[-length(exrates$USD)])
y <- log(exrates$USD[-1])
X <- matrix(c(rep(1, length(x)), x), nrow = length(x))
par(mfrow=c(1,1), mar = c(2.9, 2.9, 2.7, .5), mgp = c(1.8,.6,0), tcl = -.4)
plot(x,y, xlab=expression(log(p[t])), ylab=expression(log(p[t+1])),
    main="Scatterplot of lagged daily log prices of 1 EUR in USD",
    col="#00000022", pch=16, cex=1)
abline(0,1)
@

Irrespectively of the error specification, we expect the posterior distribution of $\betab$ to spread around $(0,1)^\top$ which corresponds to a random walk. A scatterplot of $\log p_t$ against $\log p_{t+1}$, displayed in Figure~\ref{datascatter}, confirms this.
\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=\textwidth]{Figures/jss2014-scatter}
\end{center}
\caption{Scatterplot of daily log prices at time $t$ against daily log prices at time $t+1$. The solid line indicates the identity function $f(x) = x$.}
\label{datascatter}
\end{figure}

\subsection{Posterior inference}

To obtain draws from the posterior distributions for the models with homoskedastic/SV errors, samplers developed in Section~\ref{other} are used. For the GARCH-errors, a simple random walk Metropolis-Hastings sampler is employed. We run each of the three samplers for 100\,000 iterations and discard the first 10\,000 draws as burn-in. Starting values and tuning parameters are set using maximum likelihood estimates obtained through the \proglang{R} package \pkg{fGarch} \citep{r:fGarch}.

Aiming for comparable results with minimal prior impact on predictive performance, the hyperparameters are chosen to yield vague priors: $\bb{0} = (0, 0)^\top$, $\Bb{0} = \diag{10^{10}, 10^{10}}$, $c_0 = C_0 = 0.001$, $b_\mu = 0$, $B_\mu=10^4$, $a_0=1$, $b_0 = 1$, $B_\sigmapar=1$. For the GARCH(1,1) parameter vector $\bm{\alpha} = (\alpha_0, \alpha_1, \alpha_2)^\top$, we pick independent flat priors on $\mathbb{R}^+$ for the components; the initial variance $\sigma_0^2$ is fixed to the empirical residual variance and $\tilde \y_0$ is assumed to be zero. Due to the length of the dataset (and its obvious heteroskedasticity), none of these specific choices seem to particularly influence the following analysis. Note, however, that for shorter series or series with less pronounced heteroskedasticity, sensible prior choices are of great importance as the likelihood carries less information in these cases.

The samplers yield slightly different posteriors for $\betab$, visualized in Figure~\ref{betapost}. In the top panels, the estimated marginal posterior densities $p(\beta_0|\yb)$ and $p(\beta_1|\yb)$ are displayed; the bottom panel depicts a scatterplot of draws from the joint posterior $\betab|\yb$.

<<>>=
if (usePreCalcResults) {
 load("vignette_sampling_realworld.RData")
} else {
set.seed(34567890)

#configuration parameters
draws <- 10000  #100000  TODO
thinning <- 100
burnin <- 10000
priormu <- c(-10, 1)
priorphi <- c(20, 1.5)
priorsigma <- .1
p <- 2

n <- length(x)

## design matrix:
X <- matrix(c(rep(1, n), x), nrow=n)

## prior for beta (or beta|sigma in homoskedastic regression):
b0 <- matrix(c(0, 0), nrow=ncol(X))
B0 <- diag(c(10^10, 10^10))

## prior for sigma^2 (only used in homoskedastic regression)
c0 <- 0.001
C0 <- 0.001

B0inv <- solve(B0)

# specify priors in an object
priors <- specify_priors(
  mu = sv_normal(-10, 2),
  phi = sv_beta(20, 1.5),
  sigma2 = sv_gamma(0.5, 0.5)
)

## initialize some space to hold results:
realres <- vector("list", 3)

## AR(1)-SV:
realres[[1]] <- matrix(NA_real_, nrow = floor(draws/thinning), ncol = 3 + n + p)
colnames(realres[[1]]) <- c("mu", "phi", "sigma", paste("beta", 0:(p-1), sep="_"),
			   paste("h", 1:n, sep="_"))

## some indicators:
paras <- 1:3
betas <- 3+(1:p)
latents <- 3+p+(1:n)

## starting values:
betadraw <- rep(.1, p)
paradraw <- list(mu = -10, phi = 0.8, sigma = 0.2)
latentdraw <- rep(-10, n)
paranames <- names(paradraw)

## sampler:
tim <- system.time(
for (i in -(burnin-1):draws) {
 if (i%%1000 == 0) cat("Iteration", i, "done.\n")
 # draw latent volatilities and SV-parameters:
 ytilde <- y - X %*% betadraw
 svdraw <- svsample_fast_cpp(ytilde, startpara = paradraw,
  startlatent = latentdraw, priorspec = priors)
 paradraw <- svdraw$para
 latentdraw <- drop(svdraw$latent)

 # draw the betas:
 normalizer <- as.numeric(exp(-latentdraw/2))
 Xnew <- X * normalizer
 ynew <- y * normalizer
 Sigma <- solve(crossprod(Xnew) + B0inv)
 mu <- Sigma %*% (crossprod(Xnew, ynew) + B0inv %*% b0)
 betadraw <- as.numeric(mvtnorm::rmvnorm(1, mu, Sigma))

 # store the results:
 if (i > 0 & i %% thinning == 0) {
  realres[[1]][i/thinning, paras] <- drop(paradraw)[paranames]
  realres[[1]][i/thinning, latents] <- latentdraw
  realres[[1]][i/thinning, betas] <- betadraw
 }
}
)[["elapsed"]]

## AR(1) homoskedastic:
## pre-calculate some values:
preCov <- solve(crossprod(X) + B0inv)
preMean <- preCov %*% (crossprod(X, y) + B0inv %*% b0)
preDf <- c0 + n/2 + p/2

## allocate space:
realres[[2]] <- matrix(as.numeric(NA), nrow=floor(draws/thinning), ncol=p+1)
colnames(realres[[2]]) <- c(paste("beta", 0:(p-1), sep="_"), "sigma")

## starting values:
betadraw <- rep(0, p)
sigma2draw <- var(y)

## sampler:
tim2 <- system.time(
for (i in -(burnin-1):draws) {
 if (i%%1000 == 0) cat("Iteration", i, "done.\n")
 # draw beta:
 betadraw <- as.numeric(mvtnorm::rmvnorm(1, preMean, sigma2draw*preCov))

 # draw sigma^2:
 tmp <- C0 + .5*(crossprod(y - X%*%betadraw) +
		crossprod((betadraw-b0), B0inv) %*% (betadraw-b0))
 sigma2draw <- 1/rgamma(1, preDf, rate=tmp)

 # store the results:
 if (i > 0 & i %% thinning == 0) {
  realres[[2]][i/thinning, 1:p] <- betadraw
  realres[[2]][i/thinning, p+1] <- sqrt(sigma2draw)
 }
}
)[["elapsed"]]

## AR(1)-GARCH(1,1):
## allocate space:
realres[[3]] <- matrix(NA_real_, nrow = floor(draws/thinning), ncol = 3 + n + p)
colnames(realres[[3]]) <- c("a_0", "a_1", "b_1", paste("beta", 0:(p-1), sep="_"),
			   paste("sigma", 1:n, sep="_"))

## some auxiliary functions:
loglik <- function(y, s2) {
 sum(dnorm(y, 0, sqrt(s2), log=TRUE))
}

s2s <- function(y, para, y20, s20) {
 s2 <- rep(NA_real_, length(y))
 s2[1] <- para[1] + para[2]*y20 + para[3]*s20
 for (i in 2:length(y)) s2[i] <- para[1] + para[2]*y[i-1]^2 + para[3]*s2[i-1]
 s2
}

## This is a very simple (and slow) random-walk MH sampler
## which can most certainly be improved...

## initial values:

# crude approximation of residual variance:
tmplm <- lm.fit(X, y)
s20 <- var(resid(tmplm))

y20 <- 0

## starting values:
## For the starting values for the parameters,
## use ML results from fGarch::garchFit on the
## residuals

# tmpfit <- fGarch::garchFit(~garch(1,1), resid(tmplm), trace=FALSE)
# startvals <- fGarch::coef(tmpfit)[c("omega", "alpha1", "beta1")]
startvals <- c(1.532868e-07, 3.246417e-02, 9.644149e-01)

## Random walk MH algorithm needs manual tuning for good mixing.
## Gelman et al. (1996) suggest an acceptance rate of 31.6%
## for spherical 3d multivariate normal (which we don't have :))
## These values seem to work OK:
# mhtuning <- tmpfit@fit$se.coef[c("omega", "alpha1", "beta1")] / 2
mhtuning <- c(3.463851e-08, 2.226946e-03, 2.354829e-03)

betadraw <- as.numeric(coef(tmplm))

paradraw <- startvals
paraprop <- rep(NA_real_, 3)

s2draw <- rep(.1, length(y))
s2prop <- rep(NA_real_, length(y))

accepts <- 0

## sampler:
tim3 <- system.time(
for (i in -(burnin-1):draws) {

 if (i%%1000 == 0) cat("Iteration", i, "done.\n")

 # draw volatilities and GARCH-parameters:
 ytilde <- y - X %*% betadraw  # regression residuals

 paraprop <- rnorm(3, paradraw, mhtuning)
 s2prop <- s2s(ytilde, paraprop, y20, s20)

 if (all(s2prop > 0)) { # s2 needs to be positive, otherwise reject
  logR <- loglik(ytilde, s2prop) - loglik(ytilde, s2draw)  # log acceptance rate
  if (log(runif(1)) < logR) {
   paradraw <- paraprop
   s2draw <- s2prop
   if (i > 0) accepts <- accepts + 1
  }
 }

 # draw the betas:
 normalizer <- 1/sqrt(s2draw)
 Xnew <- X * normalizer
 ynew <- y * normalizer
 Sigma <- solve(crossprod(Xnew) + B0inv)
 mu <- Sigma %*% (crossprod(Xnew, ynew) + B0inv %*% b0)
 betadraw <- as.numeric(mvtnorm::rmvnorm(1, mu, Sigma))

 # store the results:
 if (i > 0 & i %% thinning == 0) {
  realres[[3]][i/thinning, paras] <- paradraw
  realres[[3]][i/thinning, latents] <- sqrt(s2draw)
  realres[[3]][i/thinning, betas] <- betadraw
 }
}
)[["elapsed"]]

#Standardized residuals for model checking:
#homoskedastic
predmeans2 <- tcrossprod(X, realres[[2]][,1:2])
standresidmeans2 <- rowMeans((y - predmeans2)/realres[[2]][,3])

#heteroskedastic SV
predmeans <- tcrossprod(X, realres[[1]][,4:5])
standresidmeans <- rowMeans((y - predmeans)/exp(t(realres[[1]][,6:ncol(realres[[1]])])/2))

#heteroskedastic GARCH
predmeans3 <- tcrossprod(X, realres[[3]][,4:5])
standresidmeans3 <- rowMeans((y - predmeans3)/t(realres[[3]][,6:ncol(realres[[3]])]))

realresselection <- vector("list", 3)
realresselection[[1]] <- realres[[1]][,c('beta_0', 'beta_1')]
realresselection[[2]] <- realres[[2]][,c('beta_0', 'beta_1')]
realresselection[[3]] <- realres[[3]][,c('beta_0', 'beta_1')]
#save(realresselection, standresidmeans, standresidmeans2, standresidmeans3, file = 'vignette_sampling_realworld.RData')
}
@

<<betapost, fig.width=8.7, fig.height=7>>=
smootherfactor <- 1.8
par(mar = c(2.9, 2.9, 2.7, .5), mgp = c(1.8,.6,0), tcl = -.4)
layout(matrix(c(1,2,3,3), byrow=TRUE, nrow=2))
plot(density(realresselection[[1]][,"beta_0"], bw="SJ", adjust=smootherfactor),
    main=bquote(paste("p(",beta[0],"|",bold(y),")", sep="")),
    xlab="", ylab="", xlim=c(-.0005, .001))
lines(density(realresselection[[3]][,"beta_0"], bw="SJ", adjust=smootherfactor), lty=2, col=4)
lines(density(realresselection[[2]][,"beta_0"], bw="SJ", adjust=smootherfactor), lty=3, col=2)
#points(ols$coefficients[1],0, col=2)
#lines(confint(ols)[1,], rep(0,2), col=2)
legend("topleft", c("SV", "GARCH", "homosked."), col=c(1,4,2), lty=1:3)

plot(density(realresselection[[1]][,"beta_1"], bw="SJ", adjust=smootherfactor),
    main=bquote(paste("p(",beta[1],"|",bold(y),")", sep="")),
    xlab="", ylab="", xlim=c(0.9965, 1.0022))
lines(density(realresselection[[3]][,"beta_1"], bw="SJ", adjust=smootherfactor), lty=2, col=4)
lines(density(realresselection[[2]][,"beta_1"], bw="SJ", adjust=smootherfactor), lty=3, col=2)
#points(ols$coefficients[2],0, col=2)
#lines(confint(ols)[2,], rep(0,2), col=2)
legend("topright", c("SV", "GARCH", "homosked."), col=c(1,4,2), lty=1:3)

showevery <- 2L
plotorder <- sample.int(3*(nrow(realresselection[[1]]) / showevery))
cols <- rep(c("#000000bb", "#0000ff88", "#ff000088"),
	    each = nrow(realresselection[[1]]) / showevery)[plotorder]
pchs <- rep(1:3, each=nrow(realresselection[[1]]) / showevery)[plotorder]

myseq <- seq(1, nrow(realresselection[[1]]), by = showevery)
beta0 <- c(realresselection[[1]][myseq,"beta_0"],
	   realresselection[[3]][myseq,"beta_0"],
	   realresselection[[2]][myseq,"beta_0"])[plotorder]

beta1 <- c(realresselection[[1]][myseq,"beta_1"],
	   realresselection[[3]][myseq,"beta_1"],
	   realresselection[[2]][myseq,"beta_1"])[plotorder]

plot(beta0, beta1, col=cols, pch=pchs,
    xlab=bquote(paste("p(",beta[0],"|", bold(y), ")")),
    ylab=bquote(paste("p(",beta[1],"|", bold(y), ")")),
    main="Scatterplot of posterior draws")
legend("topright", c("SV", "GARCH", "homosked."), col=c(1,4,2), pch=1:3)
@

\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=\textwidth]{Figures/jss2014-betapost}
 \caption{Visualization of the posterior distributions $\betab|\yb$ for the model with SV regression residuals, the model with GARCH(1,1) regression residuals, and the model with homoskedastic regression residuals. Top panels: Kernel density estimates of the univariate posterior marginal distributions. Bottom panel: Bivariate scatterplot of posterior draws.}
  \label{betapost}
\end{center}
 \end{figure}

To assess the model fit, mean standardized residuals are depicted in Figure~\ref{qqplot}. The model with homoskedastic errors shows deficiencies in terms of pronounced dependence amongst the residuals. This can clearly be seen in the top left panel, where mean standardized residuals are plotted against time. The middle left panel shows the same plot for the model with GARCH errors where this effect is greatly diminished. The bottom left panel pictures that plot for the model with SV errors; here, this effect practically vanishes. Moreover, in the model with homoskedastic errors, the normality assumption about the unconditional error distribution is clearly violated. This can be seen by inspecting the quantile-quantile plot in the top right panel, where observed residuals exhibit much heavier tails than one would expect from a normal distribution. The model with GARCH errors provides a better fit, however, heavy tails are still visible. Standardized residuals from the model with SV errors appear to be approximately normal with only few potential outliers.

<<qqplot, fig.width=8, fig.height=7>>=
par(mfrow=c(3,2), mar = c(3.1, 3.3, 2.0, .5), mgp = c(1.7,.5,0),
    tcl = -.4)
plot(standresidmeans2, main="Residual scatterplot (homoskedastic errors)",
     xlab="Time", ylab="Standardized residuals")
qqplot(qnorm(ppoints(length(standresidmeans2)), mean=0, sd=1),
       standresidmeans2, main="Residual Q-Q plot (homoskedastic errors)",
       xlab = "Theoretical N(0,1)-quantiles", ylab = "Empirical Quantiles")
abline(0,1)
plot(standresidmeans3, main="Residual scatterplot (GARCH errors)",
     xlab="Time", ylab="Standardized residuals")
qqplot(qnorm(ppoints(length(standresidmeans3)), mean=0, sd=1),
       standresidmeans3, main="Residual Q-Q plot (GARCH errors)",
       xlab = "Theoretical N(0,1)-quantiles", ylab = "Empirical Quantiles")
abline(0,1)
plot(standresidmeans, main="Residual scatterplot (SV errors)",
     xlab="Time", ylab="Standardized residuals")
qqplot(qnorm(ppoints(length(standresidmeans)), mean=0, sd=1),
       standresidmeans, main="Residual Q-Q plot (SV errors)",
       xlab = "Theoretical N(0,1)-quantiles", ylab = "Empirical Quantiles")
abline(0,1)
@

\begin{figure}[!ht]
 \begin{center}
  \includegraphics[width=\textwidth]{Figures/jss2014-qqplot}
 \end{center}
 \caption{Visualization of mean standardized residuals. Top left panel shows a scatterplot against time for the model with homoskedastic errors, middle left panel shows this plot for the model with GARCH errors, and bottom left panel shows this plot for the model with SV errors. Quantile-Quantile plots of empirical quantiles against expected quantiles from a $\Normal{0,1}$-distribution are displayed on the panels on the right-hand side.}
 \label{qqplot}
\end{figure}

\subsection{Measuring predictive performance and model fit}
Within a Bayesian framework, a natural way of assessing the predictive performance of a given model is through its \emph{predictive density} (sometimes also referred to as \emph{posterior predictive distribution}). It is given through
\begin{eqnarray}
 \label{preddens}
p(\y_{t+1}|\yb^o_{[1:t]}) = \int\limits_{\bm K} \!
 p(y_{t+1}|\yb^o_{[1:t]}, \unobs) \times
 p(\unobs|\yb^o_{[1:t]})
 \, \dif \unobs,
\end{eqnarray}
where $\unobs$ denotes the vector of all unobservables, i.e., parameters and possible latent variables.
Note that for the model with homoskedastic errors, $\unobs=(\betab,\sigmaeps)^\top$; for the model with SV errors, $\unobs=(\betab, \allpara, \hb)^\top$; in the GARCH(1,1) case, $\unobs=(\betab,\bm{\alpha},\sigma_0^2,\tilde y_0^2)^\top$. By using the superscript $o$ in $\yb^o_{[1:t]}$, we follow \cite{gew-ami:com} and denote \emph{ex post} realizations (observations) for the set of points in time $\{1,2,\dots ,t\}$ of the \emph{ex ante} random values $\yb_{[1:t]} = (\y_1, \y_2, \dots,\y_t)^\top$. Integration is carried out over ${\bm K}$ which simply stands for the space of all possible values for $\unobs$. Equation~\ref{preddens} can be viewed as the integral of the likelihood function over the joint posterior distribution of the unobservables $\unobs$. Thus, it can be interpreted as the predictive density for a future value $\y_{t+1}$ after integrating out the uncertainty about $\unobs$, conditional on the history $\yb^o_{[1:t]}$.

In the SV errors case, Equation~\ref{preddens} is a $(n+p+3)$-dimensional integral which cannot be solved analytically. Nevertheless, it may be evaluated at an arbitrary point $x$ through Monte Carlo integration,
\begin{equation}
 \label{preddensMC}
 p(x|\yb^o_{[1:t]}) \approx \frac{1}{M}\sum_{m=1}^M p(x|\yb^o_{[1:t]}, \unobs^{(m)}_{[1:t]}),
\end{equation}
%
where $\unobs^{(m)}_{[1:t]}$ stands for the $m$\textsuperscript{th} draw from the respective posterior distribution up to time $t$. If Equation~\ref{preddensMC} is evaluated at $x=y_{t+1}^o$, we refer to it as the (one-step-ahead) \emph{predictive likelihood} at time $t+1$, denoted $\PL_{t+1}$.
Moreover, draws from the posterior predictive distribution can be obtained by simulating values $\y_{t+1}^{(m)}$ from the distribution given through the density $p(\y_{t+1}|\yb^o_{[1:t]}, \unobs^{(m)}_{[1:t]})$, the summands of Equation~\ref{preddensMC}.

For the model at hand, the predictive density and likelihood can thus be computed through the following

\begin{alg}[Predictive density and likelihood evaluation at time $t+1$]
 \label{predlikalg}
 \
 \begin{enumerate}
  \item Reduce the dataset to a training set $\yb^o_{[1:t]} = (\y_1^o,\dots,y_t^o)^\top$.
  \item Run the posterior sampler using data from the training set only to obtain $M$ posterior draws $\unobs^{(m)}_{[1:t]}$, $m=1,\dots, M$.
  \item[(3a.)] Needed for the SV model only: Simulate $M$ values from the conditional distribution $\hsv_{t+1,[1:t]}|\yb^o_{[1:t]},\unobs_{[1:t]}$ by drawing $\hsv_{t+1,[1:t]}^{(m)}$ from a normal distribution with mean $\mupar_{[1:t]}^{(m)} + \phipar_{[1:t]}^{(m)}(\hsv^{(m)}_{t,[1:t]}-\mupar_{[1:t]}^{(m)})$ and standard deviation $\sigmaparextrasub{[1:t]}^{(m)}$ for $m=1,\dots,M$.
  \item[(3b.)] Needed for the GARCH model only: Obtain $M$ draws from the conditional distribution $\sigma_{t+1,[1:t]}|\yb^o_{[1:t]},\unobs_{[1:t]}$ by computing $\sigma_{t+1,[1:t]}^{(m)} = \sqrt{\alpha_{0,[1:t]}^{(m)}+\alpha_{1,[1:t]}^{(m)}\left(\tilde y_t^o\right)^2+\alpha_{2,[1:t]}^{(m)}\left(\sigma_{t,[1:t]}^{(m)}\right)^2}$ for $m=1,\dots,M$.
  \item[4a.] To obtain $\PL_{t+1}$, average over $M$ densities of normal distributions with mean $(1,\y^o_t)\times\betab^{(m)}_{[1:t]}$ and standard deviation $\exp\{\hsv_{t+1,[1:t]}^{(m)}/2\}$ (SV model), $\sigma_{t+1,[1:t]}^{(m)}$ (GARCH model), or $\sigmaepsextrasub{[1:t]}^{(m)}$ (homoskedastic model), each evaluated at $y^o_{t+1}$, for $m=1,\dots, M$.
  \item[4b.] To obtain $M$ draws from the predictive distribution, simulate from a normal distribution with mean $(1,\y^o_t)\times\betab^{(m)}_{[1:t]}$ and standard deviation $\exp\{\hsv_{t+1,[1:t]}^{(m)}/2\}$ (SV model), $\sigma_{t+1,[1:t]}^{(m)}$ (GARCH model), or $\sigmaepsextrasub{[1:t]}^{(m)}$ (homoskedastic model) for $m=1,\dots,M$.
 \end{enumerate}
\end{alg}

It is worth pointing out that log predictive likelihoods also carry an intrinsic connection to the log \emph{marginal likelihood}, defined through
\begin{eqnarray*}
 \label{ML}
\log \ML = \log p(\yb^o) = \log \int\limits_{\bm K} \!
 p(\yb^o| \unobs) \times
 p(\unobs)
 \, \dif \unobs.
\end{eqnarray*}
This real number corresponds to the logarithm of the normalizing constant which appears in the denominator of Bayes' law and is often
used for evaluating model evidence. It can straightforwardly be decomposed into the sum of the logarithms of the one-step-ahead predictive likelihoods:
\[
\log \ML = \log p(\yb^o) = \log \prod_{t=1}^n p(\y_t^o|\yb_{[1:t-1]}^o) = \sum_{t=1}^n \log \PL_{t}.
\]
Thus, Algorithm~\ref{predlikalg} provides a conceptually simple way of computing the marginal likelihood. However, these computations are quite costly in terms of CPU time, as they require an individual model fit for each of the $n$ points in time. On the other hand, due to the embarrassingly parallel nature of the task and because of today's comparably easy access to distributed computing environments, this burden becomes manageable. E.g., the computations for the stochastic volatility analysis in this paper have been conducted in less than one hour, using 25 IBM dx360M3 nodes within a cluster of workstations. Implementation in \proglang{R} was achieved through the packages \pkg{parallel} \citep{r:r} and \pkg{snow} \citep{r:sno}.

Cumulative sums of $\log \PL_t$ also allow for model comparison through cumulative log predictive Bayes factors. Letting $\PL_t(A)$ denote the predictive likelihood of model $A$ at time $t$, and $\PL_t(B)$ the corresponding value of model $B$, the cumulative log predictive Bayes factor at time $u$ (and starting point $s$) in favor of model $A$ over model $B$ is straightforwardly given through
\begin{equation}
 \label{clpbf}
\log \left[ \frac{p_A(\yb^o_{[s+1:u]}|\yb^o_{[1:s]})}{p_B(\yb^o_{[s+1:u]}|\yb^o_{[1:s]})} \right] =  \sum_{t=s+1}^u \log \left[ \frac{\PL_t(A)}{\PL_t(B)} \right] =  \sum_{t=s+1}^u [\log \PL_t(A) - \log \PL_t(B)].
\end{equation}

When the cumulative log predictive Bayes factor is positive at a given point in time, there is evidence in favor of model $A$, and vice versa. In this context, information up to time $s$ is regarded as prior information, while out-of-sample predictive evaluation starts at time $s+1$. Note that the usual (overall) log Bayes factor is a special case of Equation~\ref{clpbf} for $s=0$ and $u=n$.

\subsection{Results}

In order to reduce prior influence, the first 1000 days are used as a training set only and the evaluation of the predictive distribution starts at $t=1001$, corresponding to December 4, 2003. The homoskedastic model is compared to the model with SV residuals in Figure~\ref{predlik1}. In the top panel, the observed series along with the 98\% one-day-ahead predictive intervals are displayed. The bottom panel shows the log one-day-ahead predictive likelihood. According to these values, SV errors can handle the inflated volatility during that time substantially better. In the course of 2009, the width of the intervals as well as the predictive likelihoods consolidate again. Figure~\ref{predlik1} also contains a close-up of the one-year time span from September 2008 to August 2009. Both credible intervals are similar at the beginning and at the end of this interval. However, there is a substantial difference in early 2009, where SV intervals become around twice as large compared to the corresponding homoskedastic analogs.

A visually barely distinguishable picture emerges when GARCH(1,1) errors are employed instead of SV errors, thus no separate figure is included in this article.

\begin{figure}[!htp]
\begin{center}
 \includegraphics[width=.984\textwidth]{Figures/jss2014-predlik1}
\end{center}
\caption{Top panel: Observed series (green) and symmetrical 98\% one-day-ahead predictive intervals for the model with homoskedastic errors (red) and the model with SV errors (black). The display also contains a close-up, showing only the period from September 2008 until August 2009. This time span is chosen to include the most noticeable ramifications of the financial crisis. During that phase, predictive performance of the model with homoskedastic errors deteriorates substantially, while SV errors can capture the inflated volatility much better. Bottom panel: Log one-day-ahead predictive likelihoods for both models.}
\label{predlik1}
\end{figure}

\begin{figure}[!htp]
\begin{center}
 \includegraphics[width=.984\textwidth]{Figures/jss2014-predlik3}
\end{center}
\caption{Top panel: Observed residuals with respect to the median of the one-day-ahead predictive distribution along with 1\% and 99\% quantiles of the respective predictive distributions. It can clearly be seen that the variances of the predictive distributions in the GARCH and SV models adjust to heteroskedasticity, while the model with homoskedastic errors is much more restrictive. Bottom panel: Cumulative log predictive Bayes factors in favor of the model with SV residuals. Values greater than zero mean that the model with GARCH/SV residuals performs better out of sample up to this point in time.}
\label{predlik3}
\end{figure}

The top panel of Figure~\ref{predlik3} visualizes the performance of the three models; it depicts observed regression residuals against their (one-day-ahead) predicted distributions. For the purpose of this image, observed regression residuals are simply defined as the deviation of the data from the median of the predicted distribution. It stands out that predictive quantiles arising from the models with heteroskedastic errors exhibit much more flexibility to adapt to the ``current state of the world'', while the simpler homoskedastic model barely exhibits this feature.

Generally speaking, there is little difference in predicted residuals until the beginning of 2007. However, during the pre-crisis era (less volatility) and during the actual crisis (more volatility), the models catering for heteroskedasticity perform substantially better. It is interesting to see that predictive intervals for the models with SV errors and GARCH errors are very similar.

In the bottom panel of Figure~\ref{predlik3}, the cumulative sums of the log predictive Bayes factors are displayed. The last points plotted equal to $176.52$ in the SV case and $166.37$ in the GARCH case; this provides overwhelming overall evidence in favor of a model catering for heteroskedasticity and ``decisive'' \citep{jef:the3ed} respectively ``very strong'' \citep{kas-raf:bay} evidence in favor of SV over vanilla GARCH(1,1) with a final cumulative predictive Bayes factor around $25\,000:1$.
%and thus clearly rejects the assumption of homoskedastic error terms for this dataset.

It is worth noting that for other exchange rates contained in \code{exrates}, a qualitatively similar picture emerges. For an overview of cumulative log predictive likelihoods, see Table~\ref{exrates}.

\begin{table}[!ht]
 \begin{small}
 \centering
 \begin{tabular}{lccc}
  \hline
  \hline
 Currency & SV & GARCH & homoskedastic \\
  \hline
  AUD & 7938 & 7890 & 7554 \\
  CAD & 7851 & 7844 & 7714 \\
  CHF & 9411 & 9337 & 8303 \\
  CZK & 9046 & 8995 & 8659 \\
  DKK\tablefootnote{For exchange rates of the Danish krone with respect to the euro, we analyze \code{1000*log(exrates\$DKK)}, i.e., per mille (\textperthousand) log prices. This way, we avoid convergence issues when obtaining starting values for the MCMC sampler from \pkg{fGarch} that appear otherwise because the krone is pegged very closely to the euro.}
  & 1473 & 1366 & 1178 \\
  GBP & 8568 & 8549 & 8218 \\
  HKD & 7907 & 7897 & 7728 \\
  IDR & 7697 & 7662 & 7269 \\
  JPY & 7607 & 7586 & 7280 \\
  KRW & 7766 & 7749 & 7188 \\
  MXN & 7536 & 7507 & 7055 \\
  MYR & 8082 & 8064 & 7928 \\
  NOK & 8648 & 8616 & 8331 \\
  NZD & 7619 & 7587 & 7440 \\
  PHP & 7890 & 7862 & 7654 \\
  PLN & 8126 & 8094 & 7727 \\
  RON & 9011 & 8880 & 8255 \\
  RUB & 8664 & 8617 & 8146 \\
  SEK & 9110 & 9101 & 8648 \\
  SGD & 8540 & 8529 & 8308 \\
  THB & 7867 & 7844 & 7692 \\
  \hline
  USD & 7878 & 7868 & 7701 \\
  USD [AR(0)]\tablefootnote{These results refer to the analysis of log returns by means of a intercept-only model, i.e., AR(0).}
  & 7876 & 7865 & 7699 \\
  \hline
  \hline
 \end{tabular}
 \caption{Final cumulative predictive log likelihoods for AR(1) models with different error assumptions, applied to the logarithm of several EUR exchange rates. % included in \pkg{stochvol}'s dataset \code{exrates}.
 SV is strongly favored in all cases.}
 \label{exrates}
 \end{small}
\end{table}

As pointed out above, the analysis of log returns can be viewed upon as a special case of the analysis of log levels where $\beta_1 \equiv 1$ is fixed a priori. When doing so, evidence in favor of heteroskedasticity is again striking. Once more, the model with SV scores highest; its demeaned predictive quantiles are almost indistinguishable from the ``varying $\beta_1$ case''. Predictive quantiles of the model with GARCH residuals again resemble those of the model with SV residuals very closely.
%Also, predictive performance is similar; the cumulative log predictive Bayes factors amount to $177.36$ for SV vs.\ homoskedastic and $166.64$ for GARCH vs.\ homoskedastic at $t=n$.
The sum of log predictive likelihoods for $t \in \{1001,1002,\dots,n\}$ is given in the last line of Table~\ref{exrates}. The values are slightly but consistently lower than for the models where $\beta_1$ is estimated from the data. The display of predictive intervals for the log returns is practically identical to Figure~\ref{predlik3} and thus omitted here.
%Figure~\ref{predlik3a} displays the predictive intervals for the log returns and the comparison through predictive likelihoods and predictive Bayes factors. It is practically identical to Figure~\ref{predlik3}.

Concluding this somewhat anecdotal prediction exercise, we would like to note that for a thorough and more universal statement concerning the real-world applicability and predictive accuracy of \pkg{stochvol}, further studies with different datasets and a richer variety of competing models, potentially including realized volatility measures, are called for. Such a voluminous undertaking is, however, beyond the scope of this paper.

\section{Conclusion}
The aim of this article was to introduce the reader to the functionality of \pkg{stochvol}. This \proglang{R} package provides a fully Bayesian simulation-based approach for statistical inference in stochastic volatility models. The typical application workflow of \pkg{stochvol} was illustrated through the analysis of exchange rate data contained in the package's \code{exrates} dataset. Furthermore, it was shown how the package can be used as a ``plug-in'' tool for other MCMC samplers. This was illustrated by estimating a Bayesian linear model with SV errors.

In the predictive example, both log levels of exchange rates from EUR to USD and log returns thereof were analyzed. For this dataset, out-of-sample analysis through cumulative predictive Bayes factors clearly showed that modeling regression residuals heteroskedastically substantially improves predictive performance, especially in turbulent times. A direct comparison of SV and vanilla GARCH(1,1) indicated that the former performs better in terms of predictive accuracy.

\section*{Acknowledgments}

The author would like to thank Sylvia Fr\"{u}hwirth-Schnatter, Hedibert Freitas Lopes, Karin Dobernig, and two anonymous referees for helpful comments and suggestions.

\bibliography{ref1}
\end{document}

### 2021-05-07 article2.Rnw
add old vignette

\documentclass[article,nojss]{jss}

%\VignetteIndexEntry{Modeling Univariate and Multivariate Stochastic Volatility in R with stochvol and factorstochvol}
%\VignetteKeyword{Bayesian inference}
%\VignetteKeyword{state-space model}
%\VignetteKeyword{heteroskedasticity}
%\VignetteKeyword{dynamic correlation}
%\VignetteKeyword{dynamic covariance}
%\VignetteKeyword{factor stochastic volatility}
%\VignetteKeyword{Markov chain Monte Carlo (MCMC)}
%\VignetteKeyword{leverage effect}
%\VignetteKeyword{asymmetric return distribution}
%\VignetteKeyword{heavy tails}
%\VignetteKeyword{financial time series}
%\VignetteEngine{knitr::knitr}
%\VignetteEncoding{UTF-8}

%% recommended packages
\usepackage{thumbpdf,lmodern}

%% our extra packages
\usepackage{bm}  % bold math
\usepackage{amsmath,amssymb}  % AMS environments

%% new custom commands
\usepackage{xparse}  % smarter commands

\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}

\newcommand{\dimy}{m}
\newcommand{\leny}{n}
\newcommand{\nfac}{r}
\newcommand{\nreg}{K}

\newcommand{\stochvol}{\pkg{stochvol}}
\newcommand{\factorstochvol}{\pkg{factorstochvol}}

\def\yvec{\bm{y}}
\def\fvec{\bm{f}}
\def\hvec{\bm{h}}

\newcommand{\phipar}{\varphi}
\newcommand{\mupar}{\mu}
\newcommand{\sigmapar}{\sigma}
\newcommand{\rhopar}{\rho}
\newcommand{\nupar}{\nu}
\newcommand{\hpar}{h}
\NewDocumentCommand{\Sigmay}{o}{\bm{\IfValueTF{#1}{#1}{}\Sigma}}  % usage: \Sigmay or \Sigmay[\tilde]  (tilde is also bold face then!)
\newcommand{\deffsvpars}[2]{%
  \expandafter\def\csname phi#1\endcsname{#2{\phipar}}
  \expandafter\def\csname mu#1\endcsname{#2{\mupar}}
  \expandafter\def\csname sigma#1\endcsname{#2{\sigmapar}}
  \expandafter\def\csname rho#1\endcsname{#2{\rhopar}}
  \expandafter\def\csname h#1\endcsname{#2{\hpar}}
  \expandafter\def\csname Sigma#1\endcsname{\Sigmay[#2]}
}
\deffsvpars{idi}{\bar}   % define all at once
\deffsvpars{fac}{\tilde} % define all at once
% now we have \phiidi, \phifac, ..., \hidi, \hfac, ...

% betaprior{fs}: f for first, s for second
\newcommand{\betapriorf}{\bm{b_\beta}}
\newcommand{\betapriors}{\bm{B_\beta}}
\newcommand{\mupriorf}{b_\mupar}
\newcommand{\mupriors}{B_\mupar}
\newcommand{\phipriorf}{a_\phipar}
\newcommand{\phipriors}{b_\phipar}
\newcommand{\rhopriorf}{a_\rhopar}
\newcommand{\rhopriors}{b_\rhopar}
\newcommand{\nuprior}{\lambda_\nupar}
\newcommand{\sigmaprior}{B_\sigmapar}

\newcommand{\Sigmalow}{\Sigmay[\check]}
\newcommand{\Loading}{\Lambda}
\newcommand{\Loadings}{\bm{\Lambda}}

\newcommand{\Real}{\mathbb{R}}
\newcommand{\Integer}{\mathbb{Z}}
\newcommand{\Student}[3]{t_{#1}(#2,#3)}   % \Student{df}{mean}{scale}
\newcommand{\Uniform}[2]{\mathcal{U}(#1,#2)}   % \Uniform{a}{b}
\newcommand{\Exponential}[1]{\mathcal{E}(#1)}   % \Exponential{a}
\newcommand{\Betadist}[2]{\mathcal{B}(#1,#2)}   % \Betadist{a}{b}
\newcommand{\Gammadist}[2]{\mathcal{G}(#1,#2)}   % \Gammadist{a}{b}
\NewDocumentCommand{\Normal}{o o m m}{\mathcal{N}\IfValueTF{#1}{_{#1}}{}\IfValueTF{#2}{\left}{}({#3},{#4}\IfValueTF{#2}{\right}{})}
% usage: \Normal{mean}{var} or \Normal[dim]{mean}{var} or \Normal[dim][1]{mean}{var}. Only the last one does \left( \right)

\newcommand{\Permmat}{\bm{P}}

\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\diag}{diag}

\def\svpars{\bm{\vartheta}}
\def\svlpars{\bm{\zeta}}

\graphicspath{{Figures/}}

\author{Darjus Hosszejni\\[.3em]WU Vienna University\\[-.1em]of Economics and Business
\And Gregor Kastner\\[.3em]University of Klagenfurt}
\Plainauthor{Darjus Hosszejni, Gregor Kastner}

\title{Modeling Univariate and Multivariate Stochastic Volatility in \proglang{R} with \stochvol{} and \factorstochvol}
\Plaintitle{Modeling Univariate and Multivariate Stochastic Volatility in R with stochvol and factorstochvol}
\Shorttitle{Modeling Stochastic Volatility in \proglang{R}}

\Abstract{
Stochastic volatility (SV) models are nonlinear state-space models that enjoy increasing popularity for fitting and predicting heteroskedastic time series.
However, due to the large number of latent quantities, their efficient estimation is non-trivial and software that allows to easily fit SV models to data is rare.
  We aim to alleviate this issue by presenting novel implementations of five SV models delivered in two \proglang{R} packages.
  Several unique features are included and documented.
  As opposed to previous versions, \stochvol{} is now capable of handling linear mean models, conditionally heavy tails, and the leverage effect in combination with SV.
  Moreover, we newly introduce \factorstochvol{} which caters for multivariate SV.
  Both packages offer a user-friendly interface through the conventional \proglang{R} generics and a range of tailor-made methods.
  Computational efficiency is achieved via interfacing \proglang{R} to \proglang{C++} and doing the heavy work in the latter.
  In the paper at hand, we provide a detailed discussion on Bayesian SV estimation and showcase the use of the new software through various examples.
}

\Keywords{%
Bayesian inference,
state-space model,
heteroskedasticity,
dynamic correlation,
dynamic covariance,
factor stochastic volatility,
Markov chain Monte Carlo (MCMC),
leverage effect,
asymmetric return distribution,
heavy tails,
financial time series%
}
\Plainkeywords{%
Bayesian inference,
state-space model,
heteroskedasticity,
dynamic correlation,
dynamic covariance,
factor stochastic volatility,
Markov chain Monte Carlo (MCMC),
leverage effect,
asymmetric return distribution,
heavy tails,
financial time series%
}

\Address{
  Darjus Hosszejni\\
  Institute for Statistics and Mathematics\\
  Department of Finance, Accounting and Statistics\\
  WU Vienna University of Economics and Business\\
  Welthandelsplatz~1 / Building D4 / Level 4\\
  1020 Vienna, Austria\\
  E-mail: \email{Darjus.Hosszejni@wu.ac.at}\\
  URL: \url{http://statmath.wu.ac.at/~hosszejni/}
  \\\\
  Gregor Kastner\\
  Department of Statistics\\
  University of Klagenfurt\\
  Universit\"atsstraße 65-67\\
  9020 Klagenfurt, Austria\\
  E-mail: \email{Gregor.Kastner@aau.at}\\
  URL: \url{http://statmath.wu.ac.at/~kastner/}
}

\begin{document}

<<setup, include=FALSE, cache=FALSE>>=
knitr::render_sweave()
knitr::opts_chunk$set(prompt = TRUE,
                      fig.show = "hide",
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE,
                      echo = FALSE,
                      cache = FALSE,
                      fig.path = "Figures/jss3813-",
                      tidy = FALSE)
base::options(continue = "+  ", prompt = "R> ", width = 70,
  useFancyQuotes = FALSE)
@

\section*{Preface}
This vignette corresponds to the forthcoming article of the same name~\citep{thisjss}.
The version at hand might receive minor updates as time goes by.
Further information about citing \pkg{stochvol} can be obtained in \proglang{R} by installing the package, e.g., through \code{install.packages("stochvol")}, and calling \code{citation("stochvol")}.

\section[Introduction]{Introduction} \label{sec:intro}

Time dependent variance is an indispensable ingredient of financial and economic time series modeling.
Already~\citet{markowitz1952portfolio} concerns himself with methods that take into account heteroskedasticity in a better way than a rolling window estimation.
By 1982, two fundamentally different approaches had been developed to cater to these needs.
On the one hand, \citet{engle1982arch} lays the groundwork for a family of time varying volatility models, most notably the generalized autoregressive conditional heteroskedasticity model~\citep[GARCH,][]{bollerslev1986garch}.
These models feature conditionally deterministic changes in the variance.
\citet{taylor1982sv}, on the other hand, addresses heteroskedasticity in his seminal work with a non-linear latent state space model, later coined the stochastic volatility (SV) model.
There, the volatility process evolves in a stochastic manner.
Despite some empirical evidence in favor of SV models over their corresponding GARCH counterparts~\citep{jacquier1994bayesian,ghysels1996stochastic,kim1998stochastic,nakajima2012bayesian}, SV and its variants enjoy little publicity among practitioners.
As~\citet{bos2012relating} underlines, one reason for this might be the lack of standard software.
In response, \citet{kastner2016dealing} provides a first version of the \proglang{R}~\citep{rlanguage} package \stochvol{} but fails to feature conditional non-Gaussianity, asymmetry (the so-called leverage effect), and multivariate generalizations.

We address these shortcomings in the manuscript at hand.
First, we extend \stochvol{}~\citep{kastner2020stochvol} with several practically relevant univariate methods.
Second, we introduce the new companion package \factorstochvol~\citep{factorstochvol} which focuses on the multivariate case.
The extended \stochvol{} now provides the means for the Bayesian estimation of vanilla SV, heavy-tailed SV, SV with leverage, and heavy-tailed SV with leverage~\citep{harvey1996estimation,omori2007stochastic,nakajima2012stochastic}.
Moreover, the package also handles these models naturally when embedded into a linear model or an autoregressive (AR) context.
The \factorstochvol{} package implements an efficient method for the Bayesian estimation of the factor SV model~\citep{kastner2017efficient}.
Among other features, the package provides several automatic factor identification schemes, hierarchical shrinkage priors~\citep[variations of the normal gamma prior,][]{griffin2010inference}, and an array of intuitive visualization methods for the high-dimensional posteriors.

The remainder of this paper is structured as follows.
We formally introduce the univariate and the multivariate models in Sections~\ref{sec:svmodels} and~\ref{sec:fsv}, respectively, including a discussion about prior distributions and a brief overview of the estimation methods.
In Section~\ref{sec:stochvol}, we unveil the new samplers of the \stochvol{} package through three example models.
We describe the \factorstochvol{} package in Section~\ref{sec:factorstochvol}, and then we conclude.

\section{Univariate SV models}\label{sec:svmodels}

We begin by introducing the vanilla SV model with linear regressors, henceforth simply called the SV model.
This is a minor but important extension of the SV model without regressors.
We also settle the notation and establish a baseline model that we generalize and reuse throughout the manuscript.
Consequently, we proceed with three generalized models: the SV model with Student's $t$~errors (SVt), the SV model with leverage (SVl), and their combination, the SV model with Student's $t$~errors and leverage (SVtl).
Finally, we close the section after discussing prior distributions and Markov chain Monte Carlo (MCMC) sampling.

\subsection{Model specifications}\label{sec:unimodels}

The key feature of the SV model is its stochastic and time-varying specification of the variance evolution.
In particular, the log-variance is assumed to follow an AR(1) process.
This feature unites the following models.

\subsubsection{Vanilla SV with linear regressors}

Let $\yvec=(y_1,\dots,y_\leny)^\top$ denote a vector of observations.
The SV model assumes the following structure for $\yvec$,
\begin{equation}\label{eq:vanillasv}
  \begin{split}
    y_t &= \bm x_t\bm\beta + \exp(\hpar_t/2)\varepsilon_t, \\
    \hpar_{t+1} &= \mupar+\phipar(\hpar_t-\mupar) + \sigmapar\eta_t, \\
    \varepsilon_t &\sim \Normal{0}{1}, \\
    \eta_t &\sim \Normal{0}{1},
  \end{split}
\end{equation}
where $\Normal{b}{B}$ denotes the normal distribution with mean $b\in\mathbb{R}$ and variance $B\in\mathbb{R}^+$, and $\varepsilon_t$ and $\eta_t$ are independent.
The log-variance process $\hvec=(\hpar_1,\dots,\hpar_\leny)^\top$ is initialized by $\hpar_0\sim\Normal{\mupar}{\sigmapar^2/(1-\phipar^2)}$.
$\bm X=(\bm x_1^\top,\dots,\bm x_\leny^\top)^\top$ is an $\leny\times\nreg$ matrix containing in its $t$th row the vector of $\nreg$ regressors at time $t$.
The $\nreg$ regression coefficients are collected in $\bm\beta=(\beta_1,\dots,\beta_\nreg)^\top$.
We refer to $\svpars=(\mupar,\phipar,\sigmapar)$ as the SV parameters: $\mupar$ is the level, $\phipar$ is the persistence, and $\sigmapar$ (also called \emph{volvol}) is the standard deviation of the log-variance.

\subsubsection{SV with Student's t errors}

Several authors have suggested to use non-normal conditional residual distributions for stochastic volatility modeling.
Examples include the Student's $t$~distribution \citep{harvey1994multivariate}, the extended generalized inverse Gaussian \citep{silva2006extented}, (semi-)parametric residuals \citep{jensen2010bayesian,delatola2011bayesian}, or the generalized hyperbolic skew Student's $t$~distribution \citep{nakajima2012stochastic}.
We implement Student's $t$~errors for the observation equation in \stochvol{}.
Formally,
\begin{equation}\label{eq:heavysv}
  \begin{split}
    y_t &= \bm x_t\bm\beta + \exp(\hpar_t/2)\varepsilon_t, \\
    \hpar_{t+1} &= \mupar+\phipar(\hpar_t-\mupar) + \sigmapar\eta_t, \\
    \varepsilon_t &\sim \Student{\nupar}{0}{1}, \\
    \eta_t &\sim \Normal{0}{1},
  \end{split}
\end{equation}
where $\varepsilon_t$ and $\eta_t$ are independent.
$\Student{\nupar}{a}{b}$ is the Student's $t$~distribution with $\nupar$ degrees of freedom, mean $a$, and variance $b$.
The single difference between Equation~\ref{eq:vanillasv} and Equation~\ref{eq:heavysv} is that here the observations are conditionally $t$~distributed.
Hence, Equation~\ref{eq:heavysv} generalizes Equation~\ref{eq:vanillasv} through the new parameter $\nupar$ as the Student's $t$~distribution converges in law to the standard normal distribution when $\nupar$ goes to infinity.

\subsubsection{SV with leverage}

Propositions for asymmetric innovations include non-parametric distributions~\citep{jensen2014estimating}, skewed distributions~\citep{nakajima2012stochastic}, and distributions featuring correlation with the variance process, also called the leverage effect~\citep{harvey1996estimation,jacquier2004bayesian}.
We implement the leverage effect in the \stochvol{} package.
Formally,
\begin{equation}\label{eq:svleverage}
  \begin{split}
    y_t &= \bm x_t\bm\beta + \exp(\hpar_t/2)\varepsilon_t, \\
    \hpar_{t+1} &= \mupar+\phipar(\hpar_t-\mupar) + \sigmapar\eta_t, \\
    \varepsilon_t &\sim \Normal{0}{1}, \\
    \eta_t &\sim \Normal{0}{1},
  \end{split}
\end{equation}
where the correlation matrix of $(\varepsilon_t,\eta_t)$ is
\begin{equation}\label{eq:corrmatrix}
  \bm\Sigma^\rhopar = \begin{pmatrix} 1 && \rhopar \\ \rhopar && 1 \end{pmatrix}.
\end{equation}
The vector $\svlpars=(\mupar,\phipar,\sigmapar,\rhopar)^\top$ collects the SV parameters.
The new parameter compared to Equation~\ref{eq:vanillasv} is a correlation term $\rhopar$ which relates the residuals of the observations to the innovations of the variance process.
Equation~\ref{eq:vanillasv} is therefore a special case of Equation~\ref{eq:svleverage} with a pre-fixed $\rhopar=0$.

\subsubsection{SV with Student's t errors and leverage}

Some authors have proposed the combination of $t$~errors with the leverage effect~\citep{jacquier2004bayesian,omori2007stochastic,nakajima2009leverage}.
We implement the common generalization of Equation~\ref{eq:heavysv} and Equation~\ref{eq:svleverage}.
Formally,
\begin{equation}\label{eq:heavysvleverage}
  \begin{split}
    y_t &= \bm x_t\bm\beta + \exp(\hpar_t/2)\varepsilon_t, \\
    \hpar_{t+1} &= \mupar+\phipar(\hpar_t-\mupar) + \sigmapar\eta_t, \\
    \varepsilon_t &\sim \Student{\nupar}{0}{1}, \\
    \eta_t &\sim \Normal{0}{1},
  \end{split}
\end{equation}
where the correlation matrix of $(\varepsilon_t,\eta_t)$ is $\bm\Sigma^\rhopar$ as in Equation~\ref{eq:corrmatrix}.

\subsection{Prior distributions}\label{sec:svpriors}

We a priori assume $\bm\beta\sim\Normal[\nreg]{\betapriorf}{\betapriors}$, where $\Normal[l]{\bm b}{\bm B}$ is the $l$-dimensional normal distribution with mean vector $\bm b$ and variance-covariance matrix $\bm B$.
For small values in the diagonal of $\betapriors$, this prior enforces shrinkage towards $\betapriorf$; for large values in the diagonal, the prior turns rather uninformative.
By setting $\betapriorf$ to the zero vector and $\betapriors$ to a scaled identity matrix, the prior distribution becomes the Bayesian analogue to ridge regression~\citep[see, e.g.,][for a discussion of this and other shrinkage priors]{park2008bayesian}.

The level $\mupar\in\Real$ is unrestricted, hence we can apply the common $\mupar\sim\Normal{\mupriorf}{\mupriors}$ prior.
Depending on the application, a fairly uninformative distribution is the usual choice, e.g., setting $\mupriorf=0$ and $\mupriors\ge100$ for daily asset log returns.
In our experience, the exact values of the prior mean and prior variance of $\mupar$ do not strongly affect the estimation results unless $\mupriors$ is small.

To achieve stationarity in the variance process, a restricted persistence $\phipar\in(-1,1)$ is needed.
To this end, we assume $(\phipar+1)/2\sim\Betadist{\phipriorf}{\phipriors}$, where $\Betadist{\phipriorf}{\phipriors}$ is the beta distribution with shape parameters $\phipriorf$ and $\phipriors$.
The selection of the shape parameters may be relatively influential with many data sets.
In financial applications with daily asset log returns, the variance tends to be highly persistent, i.e., $\phipar\approx1$.
Such domain knowledge can be used as prior information by allocating more probability to positive high values of $\phipar$, e.g., by setting $\phipriorf\gtrsim5$ and $\phipriors\approx1.5$.
As an alternative, when stationarity is not assumed, the untruncated prior $\phipar\sim\Normal{b_\phipar}{B_\phipar}$ can also be applied.

The volvol is positive but we would like allow $\sigmapar$ to approach 0 as closely as needed -- that allows us to be less informative and to improve the estimates.
Following~\citet{fruhwirth2010stochastic} and~\citet{kastner2014ancillarity}, we advocate $\sigmapar\sim\lvert\Normal{0}{\sigmaprior}\rvert$ instead, where $\lvert\Normal{0}{\sigmaprior}\rvert$ denotes the half normal distribution.
It corresponds to $\sigmapar^2\sim\Gammadist{1/2}{1/(2\sigmaprior)}$, where $\Gammadist{a}{b}$ is the gamma distribution with shape parameter $a$ and rate parameter $b$.
As an alternative, the commonly applied and convenient conjugate gamma prior on $\sigmapar^{-2}$ can be assumed.
However, it bounds $\sigmapar$ away from 0 and it is therefore in our view an unsatisfactory choice.

As a last step in fully specifying the vanilla SV model in Equation~\ref{eq:vanillasv}, the variance process is initialized a priori with its stationary distribution, i.e., $\hpar_0\sim\Normal{\mupar}{\sigmapar^2/(1-\phipar^2)}$.
This consistently extends our prior assumptions about $\hvec$ following a stationary AR(1) process.
As an alternative, when stationarity is not assumed, $\hpar_0\sim\Normal{\mupar}{B_h}$ can be applied with a constant variance $B_h$.

The SV models with Student's $t$~errors additionally require the prior specification of the degrees of freedom parameter $\nupar$.
To ascertain interpretability of the scaling $\exp(\hpar_t/2)$, we ensure finite second moments of $\yvec$ by enforcing $\nupar>2$.
As a reviewer recommended, we follow~\citep{geweke1993bayesian} and equip $\nupar$ with an exponential prior $\nupar-2\sim\Exponential{\nuprior}$, where $\nuprior$ is the rate of the exponential distribution.

Finally, in the case of the SV models with leverage, we employ the translated and scaled beta distribution for $\rhopar\in(-1,1)$ as in~\citet{omori2007stochastic}, i.e., $(\rhopar+1)/2\sim\Betadist{\rhopriorf}{\rhopriors}$.
We find that the posterior estimates of $\rhopar$ can be sensitive to its prior distribution, thus, some care is needed when setting the hyperparameters in practice.
In our experience, slightly informative choices such as $\rhopriorf=\rhopriors\approx4$ work well in financial applications.

\subsection{Estimation}

All methods implemented in \stochvol{} and \factorstochvol{} rely on the Bayesian paradigm.\footnote{At this point we would also like to point out works aiming at estimating stochastic volatility and related models within the frequentist framework, see, e.g., \citet{abanto2017maximum, creal2017class}, and in particular the recent \pkg{stochvolTMB} package \citep{stochvolTMB}.}
Bayesian analysis aims to estimate model parameters through Bayesian updating.
By using probability distributions to represent information, Bayes' theorem can be employed to update the prior information to the posterior information by incorporating the observations. This approach has the advantage of providing full uncertainty quantification in a probabilistic framework without relying on asymptotic results; moreover, so-called \emph{shrinkage priors} can be used to regularize the posterior and guard against overfitting. For an introductory textbook on Bayesian statistics, see, for instance, \citet{mcelreath2015statistical}.

When the posterior distribution is not available analytically, one customarily resorts to approximations such as perfect simulation~\citep{huber2015perfect}, approximate Bayesian computation~\citep{sisson2018handbook}, adaptive Monte Carlo methods~\citep{roberts2005coupling}, or MCMC methods.
When computationally feasible, MCMC is a valuable tool that provides draws from the posterior distribution in question.
That way, MCMC approximates the posterior distribution similarly to a histogram approximating a density.
For a more in-depth introduction on MCMC methods, see, for instance, \citet{brooks2011handbook}.

The estimation algorithm of SV, SVt, SVl, and SVtl all resemble the original methodology developed in~\citet{kastner2014ancillarity} for the vanilla SV model.
Namely, to draw from the posterior distribution of $\hvec$ efficiently, the MCMC sampler resorts to approximate mixture representations of Equations~\ref{eq:vanillasv},~\ref{eq:heavysv},~\ref{eq:svleverage} and~\ref{eq:heavysvleverage} similar to the ones in \citet{kim1998stochastic} and \citet{omori2007stochastic}.
Doing so yields a conditionally Gaussian state space model for which efficient sampling methods are available~\citep{fruhwirth1994data,carter1994gibbs}.
Following~\cite{rue2001fast} and~\cite{mccausland2011simulation}, we draw the full vector $\hvec$ ``all without a loop'' (AWOL).

When Student's $t$~errors with unknown degrees of freedom are used, we handle the added complication through the well-known representation of the $t$~distribution as a scale mixture of Gaussians.
This requires additional Gibbs and independence Metropolis-Hastings steps documented in~\citet{kastner2015heavy}.
Furthermore, we deal with the increased complexity in the posterior space of the leverage case by repeated ancillarity-sufficiency interweaving strategies~\citep[ASIS,][]{yu2011to} steps in the sampling scheme, see~\citet{hosszejni2019approaches} for details.

To verify the correctness of the implementation, unit tests are included in the package which can be run by \code{devtools::test()}~\citep{devtools}.
In particular, a variant of Geweke's test~\citep{geweke2004getting} is part of the test suite.
In this test, we exploit that the sampling distribution of the model parameters during the Geweke test is identical to their preset prior distribution.
Therefore, the cumulative distribution function maps the sample to a uniform distribution, which in turn is mapped to a normal distribution using the normal distribution's quantile function.
If the user chooses to execute the automated unit tests in \stochvol{}, the system evaluates the thinned and transformed sample using the \code{shapiro.test()} function, where the thinning of the sample is done to approximate independent sampling.

For maximal computational effectiveness, all sampling algorithms are implemented in the compiled language \proglang{C++}~\citep{cpplanguage} with the help of the \proglang{R} package \pkg{Rcpp}~\citep{eddelbuettel2011rcpp}.
Matrix computations make use of the efficient \proglang{C++} template library \pkg{Armadillo}~\citep{sanderson2016armadillo} through the \proglang{R} package \pkg{RcppArmadillo}~\citep{eddelbuettel2014rcpparmadillo}.\footnote{For explicit run time discussions please see \cite{kastner2014ancillarity} and \cite{hosszejni2019approaches}. For the possibility to use multi-core computing within a single MCMC chain and potential speed gains when doing so, please see \cite{kastner2019sparse}.}
After sampling, the resulting \proglang{R} objects make use of plotting and summary functions of the \proglang{R} package \pkg{coda}~\citep{coda}.

\section{Multivariate SV models} \label{sec:fsv}

A key difficulty accompanying dynamic covariance estimation is the relatively high number of unknowns compared to the number of observations.
More precisely, letting $\dimy$ denote the cross-sectional dimension, the corresponding covariance matrix $\Sigmay_t$ contains $\dimy(\dimy+1)/2$ degrees of freedom, a quadratic term in $\dimy$.
Table~\ref{tab:numpar1} illustrates the ``curse of dimensionality'' for various values of $\dimy$.
One way to break this curse is to use latent factors and thereby achieve a sparse representation of $\Sigmay_t$.

\begin{table}[t]
  \centering
  \begin{tabular}{lll}
    $\dimy$ & free elements of $\Sigmay_t$ & free elements of $\Sigmay_t$ per data point \\ \hline
    1       & 1                            & 1                                           \\
    10      & 55                           & 5.5                                         \\
    100     & 5050                         & 50.5                                        \\
    1000    & 500500                       & 500.5
  \end{tabular}
  \caption{Absolute and relative numbers of free elements of the time-varying covariance matrix $\Sigmay_t$ for different numbers of component series $\dimy$.}
  \label{tab:numpar1}
\end{table}

\subsection{The factor SV model}

Latent factor models embody the idea that even high dimensional systems are driven by only a few sources of randomness.
These few sources of randomness control a few factors, which in turn account for the interactions between the observations.
Moreover, latent factor models provide an efficient tool for dynamic covariance matrix estimation.
They allow for a reduction in the number of unknowns.
A conventional latent factor model with $\nfac$ factors implies the decomposition
\begin{equation}\label{eq:decomp}
  \Sigmay_t=\Sigmalow_t+\Sigmaidi_t,
\end{equation}
where $\rank(\Sigmalow_t)=\nfac<\dimy$, and $\Sigmaidi_t$ is the diagonal matrix containing the variances of the idiosyncratic errors.
The rank assumption on the symmetric $\Sigmalow_t$ gives rise to the factorization $\Sigmalow_t=\Psi\Psi^\top$, where $\Psi\in\Real^{\dimy\times\nfac}$ contains $mr-\nfac(\nfac-1)/2$ free elements~\citep[see, e.g., the pivoted Cholesky algorithm in][]{higham1990analysis}.
Hence, $\dimy(\nfac+1)-\nfac(\nfac-1)/2$ free elements remain in $\Sigmay_t$, now only linear in $\dimy$.
Table~\ref{tab:numpar2} illustrates the ``broken curse of dimensionality'' for various values of $\dimy$ and $\nfac=4$.

\begin{table}[t]
  \centering
  \begin{tabular}{lll}
    $\dimy$ & free elements of $\Sigmay_t$ & free elements of $\Sigmay_t$ per data point \\ \hline
    10      & 44                           & 4.4                                         \\
    100     & 494                          & 4.94                                        \\
    1000    & 4994                         & 4.994
  \end{tabular}
  \caption{Absolute and relative numbers of free elements of the time-varying covariance matrix $\Sigmay_t$ in a factor model for different numbers of component series $\dimy$ and number of factors $\nfac=4$.}
  \label{tab:numpar2}
\end{table}

In the following, we describe the factor SV model employed in the \factorstochvol{} package.
We model the observations $\yvec_t=(y_{t1},\dots,y_{t\dimy})^\top$ as follows.
\begin{equation}\label{eq:fsv}
  \begin{split}
    \yvec_t\mid\bm\beta,\Loadings,\fvec_t,\Sigmaidi_t &\sim \Normal[\dimy]{\bm\beta + \Loadings\fvec_t}{\Sigmaidi_t}, \\
    \fvec_t\mid\Sigmafac_t &\sim \Normal[\nfac]{\bm 0}{\Sigmafac_t},
  \end{split}
\end{equation}
where $\fvec_t=(f_{t1},\dots,f_{t\nfac})^\top$ is the vector of factors, $\bm\beta=(\beta_1,\dots,\beta_\dimy)^\top$ is an observation-specific mean, and $\Loadings\in\Real^{\dimy\times\nfac}$ is a tall matrix holding the factor loadings.
The covariance matrices $\Sigmaidi_t$ and $\Sigmafac_t$ are both diagonal representing independent vanilla SV processes.
\begin{equation}\label{eq:fsvcov}
  \begin{split}
    \Sigmaidi_t &= \diag(\exp(\hidi_{t1}),\dots,\exp(\hidi_{t\dimy})), \\
    \Sigmafac_t &= \diag(\exp(\hfac_{t1},\dots,\exp(\hfac_{t\nfac}))), \\
    \hidi_{ti} &\sim \Normal{\muidi_i+\phiidi_i(\hidi_{t-1,i}-\muidi_i)}{\sigmaidi_i^2}, \quad i=1,\dots,\dimy, \\
    \hfac_{tj} &\sim \Normal{\mufac_j+\phifac_j(\hfac_{t-1,j}-\mufac_j)}{\sigmafac_j^2}, \quad j=1,\dots,\nfac.
  \end{split}
\end{equation}
For a more theoretical treatment of factor SV from a Bayesian point of view, the reader is referred to, e.g., \citet{pitt1999time}, \citet{aguilar2000bayesian}, \citet{chib2006analysis}, and \citet{han2006asset}.

Based on Equation~\ref{eq:fsv}, we can reformulate Equation~\ref{eq:decomp} as
\begin{equation}\label{eq:vardecomp}
  \Sigmay_t=\Loadings\Sigmafac_t\Loadings^\top+\Sigmaidi_t,
\end{equation}
from which several identification issues are apparent: the order, the sign, and the scale of the factors is unidentified.
More specifically, for any generalized permutation matrix%
\footnote{A generalized permutation matrix has the zero--non-zero pattern of a permutation matrix, but it is allowed to have any non-zero values instead of just ones.
Hence, a generalized permutation matrix not only permutes but also scales and switches the sign of its multiplier.} %
$\Permmat$ of size $\nfac\times\nfac$, we find another valid decomposition $\Sigmay_t=\Loadings^\prime\Sigmafac^\prime_t(\Loadings^\prime)^\top+\Sigmaidi_t$, where $\Loadings^\prime=\Loadings\Permmat^{-1}$ and
$\Sigmafac^\prime_t=\Permmat\Sigmafac_t\Permmat^\top$.
We resolve the ambiguity in the scale of the factors by fixing the level of their log-variance to zero, i.e., $\mufac_j=0$ for $j=1,\dots,\nfac$.
Sign and order identification can be enforced through restrictions on the factor loadings matrix $\Loadings$.
Several options are available in \factorstochvol{} for restricting $\Loadings$, for details see Section~\ref{sec:fsvuse2}.

\subsection{Prior distributions}\label{sec:fsvpriors}

Priors need to be specified for the mean, the latent log-variance processes, and for the factor loadings matrix $\Loadings$.
We choose $\beta_j\sim\Normal{b_\beta}{B_\beta}$, independently for $j=1,\dots,\dimy$.
For small values of $B_\beta$, this shrinks $\beta_j$ toward $b_\beta$; for large values of $B_\beta$, the prior is fairly uninformative.

The log-variance processes have the same prior specification as in the univariate case in Section~\ref{sec:svpriors}.
For $\Loadings$, three types of priors are currently implemented in \factorstochvol.
All three can be written in the form $\Loading_{ij}\sim\Normal{0}{\tau^2_{ij}}$ independently for each applicable $i\in\{1,\dots\dimy\}$ and $j\in\{1,\dots,\nfac\}$.
First, one can fix all the $\tau^2_{ij}$s -- not necessarily to the same value -- a priori.
This results in a normal prior for each element of the loadings matrix.

The second type is a hierarchical prior which has been developed to induce more flexible and potentially stronger shrinkage,
\begin{equation} \label{eq:ngprior}
  \Loading_{ij}\mid\tau_{ij}^2 \sim \Normal{0}{\tau_{ij}^2}, \quad \tau_{ij}^2\mid\lambda_i^2 \sim \Gammadist{a}{a\lambda_i^2/2}.
\end{equation}
This distribution is termed normal gamma prior by \citet{griffin2010inference} and implies a conditional variance $\mathbb{V}(\Loading_{ij}\mid\lambda_i^2)$ of $2/\lambda_i^2$ and an unconditional excess kurtosis of $3/a$.
The value of $a$ is treated as a structural parameter to be fixed by the user, where choosing $a$ small ($\lesssim 1$) enforces strong shrinkage towards zero, while choosing $a$ large ($\gtrsim 1$) imposes little shrinkage.
The case $a = 1$ is a special case termed the Bayesian Lasso prior~\citep{park2008bayesian}.
The parameter $\lambda_i^2$ is estimated from the data with $\lambda_i^2 \sim \Gammadist{c}{d}$.

The third type is a slight modification of the second.
Because variances in each row of the factor loadings matrix $\Loadings$ can be seen as ``random effects'' from the same underlying distribution, the prior in Equation~\ref{eq:ngprior} induces row-wise shrinkage with element-wise adaption.
Analogously, one could also consider column-wise shrinkage with element-wise adaption, i.e.,
\begin{equation} \label{eq:ngpriorCW}
  \Loading_{ij}\mid\tau_{ij}^2\sim\Normal{0}{\tau_{ij}^2}, \quad \tau_{ij}^2\mid\lambda_j^2\sim\Gammadist{a}{a\lambda_j^2/2},
\end{equation}
with the corresponding prior $\lambda_j^2 \sim \Gammadist{c}{d}$.

\subsection{Estimation}

Bayesian estimation in the factor SV model builds on the univariate vanilla SV implementations in \stochvol{} and features several levels of efficiency boosting. To alleviate the problem of potentially slow convergence in high dimensions, it is carried out via a sampler that utilizes several variants of ASIS. The sampling details implemented in \factorstochvol{} are described in \citet[using Gaussian priors for the factor loadings]{kastner2017efficient} as well as \citet[using hierarchical shrinkage priors for the factor loadings]{kastner2019sparse}.

Similarly to \stochvol{} and in an attempt to make computation time bearable even in higher dimensions, \factorstochvol{}'s main sampler is written in \proglang{C++}.
It uses the \proglang{R} package \pkg{Rcpp} to ease communication between \proglang{R} and \proglang{C++}. The univariate SV parts are borrowed from \stochvol{} and interfaced through its \proglang{C}/\proglang{C++}-level updating function \fct{update\_fast\_sv}. In doing so, moving between interpreted \proglang{R} code and compiled \proglang{C++} code at each MCMC iteration is avoided.

\section[The stochvol package]{The \stochvol{} package} \label{sec:stochvol}%

The \stochvol{} package provides means for fitting univariate SV, SVt, SVl, and SVtl models via its sampling routines \fct{svsample}, \fct{svtsample}, \fct{svlsample}, and \fct{svtlsample}, respectively.
In the following, we describe a recommended workflow with \stochvol.
First, we discuss estimation, visualization, and prediction using default settings.
Then, we show how to adapt the values of the prior hyperparameters and how to configure the sampling mechanism.

\subsection{Preparing the data and running the MCMC sampler} \label{sec:svuse1}

We estimate three models that exemplify the features and the user interface of \stochvol.
Using the \code{exrates} data found in the package, we model the EURCHF exchange rate (the price of 1 euro in Swiss franc) in the period between March 1, 2008 and March 1, 2012 (1028 data points) in three different ways.

\subsubsection{AR(1) model with SV residuals}

The first example is an AR(1) model with SV residuals, i.e., Equation~\ref{eq:vanillasv} turns into
\begin{equation*}
  \begin{split}
    y_t\mid y_{t-1},\beta_0,\beta_1,\hpar_t &\sim \Normal{\beta_0+\beta_1y_{t-1}}{\exp(\hpar_t)}, \\
    \hpar_{t+1}\mid\svpars,\hpar_t &\sim \Normal{\mupar+\phipar(\hpar_t-\mupar)}{\sigmapar^2}.
  \end{split}
\end{equation*}
Using this model, we test whether the exchange rate follows a random walk with SV.
In this case, we expect the posteriors of $\beta_0$ and $\beta_1$ to concentrate around 0 and 1, respectively.

In order to estimate this AR(1)-SV model, we need to prepare the input $\yvec$ as a \code{numeric} sequence of length $\leny$ and pass it as the first input argument to \fct{svsample} as follows:
<<presvrunmodel, eval=TRUE, results='hide'>>=
set.seed(1)
library("stochvol")
data("exrates")
ind <- which(exrates$date >= as.Date("2008-03-01") &
  exrates$date <= as.Date("2012-03-01"))
CHF_price <- exrates$CHF[ind]
@
<<svrunmodel, results='hide', echo=TRUE>>=
<<presvrunmodel>>
res_sv <- svsample(CHF_price, designmatrix = "ar1")
@
We set \code{designmatrix = "ar1"} to use the AR(1) specification.
More generally, \code{designmatrix} can take \code{character} values of the form \code{"ar0"} for a constant mean model, or \code{"ar1"}, \code{"ar2"}, etc., for AR(1), AR(2), and so on.

\subsubsection{Constant mean model with SVt residuals}

The second example is a constant mean model with SVt residuals, i.e., Equation~\ref{eq:heavysv} becomes
\begin{equation*}
  \begin{split}
    y_t\mid\beta_0,\hpar_t,\nupar &\sim \Student{\nu}{\beta_0}{\exp(\hpar_t/2)}, \\
    \hpar_{t+1}\mid\svpars,\hpar_t &\sim \Normal{\mupar+\phipar(\hpar_t-\mupar)}{\sigmapar^2}.
  \end{split}
\end{equation*}
If the returns are heavy-tailed, most of the posterior mass of $\nupar$ concentrates on low values, e.g., smaller than 20.
Otherwise, there is little evidence for high kurtosis.

We compute the log returns by applying \fct{logret} on the previously calculated \code{CHF\_price}.
Then, to estimate the constant mean model with heavy tailed SV residuals, we pass the vector of log returns to \fct{svtsample} with \code{designmatrix} set to \code{"ar0"}.
<<presvtrunmodel, results='hide', eval=TRUE, dependson="presvrunmodel">>=
set.seed(2)
CHF_logret <- 100 * logret(CHF_price)
@
<<svtrunmodel, echo=TRUE, results='hide'>>=
<<presvtrunmodel>>
res_svt <- svtsample(CHF_logret, designmatrix = "ar0")
@

\subsubsection{Multiple regression with SVl residuals}

The third example is a multiple regression model with an intercept, two regressors, and SVl residuals; that is, Equation~\ref{eq:svleverage} turns into
\begin{equation*}
  \begin{split}
    \begin{pmatrix} y_t \\ \hpar_{t+1} \end{pmatrix} \;\Bigg|\; \hpar_t,\svlpars,\begin{pmatrix} x_{t1} \\ x_{t2} \end{pmatrix},\beta_0,\begin{pmatrix} \beta_1 \\ \beta_2 \end{pmatrix} &\sim
    \Normal[2][1]{\begin{pmatrix} \beta_0+\beta_1x_{t1}+\beta_2x_{t2} \\ \mupar+\phipar(\hpar_t-\mupar) \end{pmatrix}}{\bm\Sigma^\rhopar}, \\
    \bm\Sigma^\rhopar &= \begin{pmatrix} \exp(\hpar_t) && \rhopar\sigmapar\exp(\hpar_t/2) \\ \rhopar\sigmapar\exp(\hpar_t/2) && \sigmapar^2 \end{pmatrix}.
  \end{split}
\end{equation*}
For illustration, we regress EURCHF log returns onto the contemporaneous log returns on EURUSD and EURJPY, the value of 1 euro per US dollar and Japanese yen, respectively.

To estimate a multiple regression model using \stochvol, we need to prepare a \code{numeric} matrix $\bm X$ of dimension $\leny\times\nreg$, where rows correspond to time points and columns to covariates.
We create an intercept as the first column of $\bm X$, and we set the second and the third columns to the EURUSD log returns and the EURJPY log returns, respectively; finally, we use the columns of $\bm X$ as covariates in the multiple regression.
<<svlrunmodel, echo=TRUE, eval=TRUE, dependson="presvtrunmodel">>=
set.seed(3)
X <- cbind(constant = 1, 100 * logret(exrates$USD[ind]),
  100 * logret(exrates$JPY[ind]))
res_svl <- svlsample(CHF_logret, designmatrix = X)
@

\subsection{Visualizing the results} \label{sec:svuse2}

Often, the joint posterior distribution of model parameters and latent quantities mark the goal of a Bayesian analysis.
To inspect it, one can look at summary statistics and various types of visualizations of marginal posterior distributions.
Also, it is recommended to examine the Markov chain for possible convergence issues -- this happens usually by investigating trace plots of posterior quantities.
For this reason, inspired by the \pkg{coda} package, \stochvol{} provides its own instances of the \proglang{R} generic functions \fct{plot} and \fct{summary}.
In order to introduce the tools that \stochvol{} provides for analyzing MCMC output, we briefly examine the results of the third example (multiple regression with SVl errors) in the remaining part of the section.

First, we plot the output of the estimation.
<<svlplot, echo=TRUE, dependson="svlrunmodel", fig.height=4, results='hide'>>=
plot(res_svl, showobs = FALSE, dates = exrates$date[ind[-1]])
@
The result is shown in Figure~\ref{fig:svlall}.
We see in the first row a summary of the posterior density of the volatility. Apart from its median, we also receive a quantification of the uncertainty through the 5\% and the 95\% quantiles at each time point.
In the second row, we can follow the evolution of the Markov chain of the SV parameters. In this example, they are $\mupar$, $\phipar$, $\sigmapar$, and $\rhopar$.
Lastly, we see prior and posterior density plots of the parameters in the third row in gray and black, respectively.
They show high persistence and significant leverage.

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{jss3813-svlplot-1.pdf}
  \caption{The default plot of an estimated model.
  The top row shows a summary of the posterior of the daily volatility (in percent) $100\exp(\bm h/2)$ through its median (black) and 5\% and 95\% quantiles (gray).
  The remaining panels summarize the Markov chains of the parameters $\mupar$, $\phipar$, $\sigmapar$, and $\rhopar$.
  In particular, the middle row presents trace plots and the bottom row shows prior (gray, dashed) and posterior (black, solid) densities.}
  \label{fig:svlall}
\end{figure}

Next, we observe the AR coefficients.
<<svlbetaplot, echo=2, dependson="svlrunmodel", fig.width=6.7, fig.height=3.5, results='hide'>>=
par(mar = c(2.5, 1.5, 0.5, 0.5), mfrow = c(3, 2), mgp = c(1.7, 0.5, 0))
for (i in seq_len(3)) {
  coda::traceplot(svbeta(res_svl)[, i])
  coda::densplot(svbeta(res_svl)[, i], show.obs = FALSE)
}
@
The result is shown in Figure~\ref{fig:svlbeta}.
On the left hand side, we do not spot any signs of convergence or mixing problems in the trace plots.
On the right hand side, we see that none of the posterior densities of $\beta_0$, $\beta_1$, and $\beta_2$ concentrate around 0, hence the covariates seem to have an impact on the dependent variable.
\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{jss3813-svlbetaplot-1}
  \caption{Trace plots and estimated kernel densities of posterior draws from $p(\bm\beta\mid\bm y)$.}
  \label{fig:svlbeta}
\end{figure}

As the final step, we print a numeric summary of the estimation results.
<<printsummary, echo=TRUE, eval=TRUE, results="markup">>=
summary(res_svl, showlatent = FALSE)
@
For brevity, we set \code{showlatent = FALSE} in order not to print all the 1027 latent states.
The output shows the length of the burn-in and the number of draws, the prior specification of the parameters, and a concise summary of the marginal posterior distributions of the parameters $\mupar$, $\phipar$, $\sigmapar$, and $\rhopar$, and additionally of the level of the volatility $\exp(\mupar/2)$ and of $\sigmapar^2$, and of the vector of regression coefficients $\bm\beta$.
This posterior summary is a table consisting of columns for the posterior mean and standard deviation, the 5\%, 50\%, and 95\% quantiles.
The user can influence the shown quantiles by passing a sequence of values between 0 and 1 to \fct{svsample}, \fct{svtsample}, \fct{svlsample}, or \fct{svtlsample} via the argument \code{quantiles}.

The last column in the table depicts the so-called effective sample size (ESS), a measure of the quality of a converged MCMC chain.
Formally, ESS of a Markov chain $\mathrm{C}$ is defined through $M/(1+2\sum_{s=1}^\infty\rho^\text{eff}(s))$, where $M$ is the length of $\mathrm{C}$ and $\rho^\text{eff}(s)$ denotes the autocorrelation function for lag $s$ among the elements of $\mathrm{C}$.
In principle, ESS is the sample size of a serially uncorrelated chain bearing the same Monte Carlo error as our (marginal) chain.
Intuitively speaking, this means that ESS is the number of independent and identically distributed draws that were acquired and gives a sense of how well our chain has explored the posterior space.
Higher values of ESS indicate better mixing.

\subsection[Prediction with stochvol]{Prediction with \stochvol}
We employ our estimated model to predict log returns for the remaining days in the data set.
To do so, we first prepare the covariates for the next 24 days and pass them via the argument \code{newdata} of the generic \fct{predict} function along with the estimation output.
Note that we need 25 days of price data to obtain 24 returns.
<<svlpredict, echo=TRUE, eval=TRUE>>=
set.seed(4)
pred_ind <- seq(tail(ind, 1), length.out = 25)
pred_X <- cbind(constant = 1, 100 * logret(exrates$USD[pred_ind]),
  100 * logret(exrates$JPY[pred_ind]))
pred_svl <- predict(res_svl, 24, newdata = pred_X)
@
As we have access to the entire distribution of future log returns, we can quantify the uncertainty around our predictions through quantiles.
In the following code snippet, we visualize the $k$-step-ahead predictive distributions for $k=1,\dots,24$, along with the truly observed values.
The result is in Figure~\ref{fig:svlpred}.
<<plotsvlpred, echo=2:4, eval=TRUE, fig.height=3.5, fig.width=10, results='hide'>>=
par(mgp = c(1.7, 0.5, 0))
obs_CHF <- 100 * logret(exrates$CHF[pred_ind])
qs <- t(apply(predy(pred_svl), 2, quantile, c(0.05, 0.5, 0.95)))
ts.plot(cbind(qs, obs_CHF), xlab = "Periods ahead", lty = c(rep(1, 3), 2),
  col = c("gray80", "black", "gray80", "red"))
@
\begin{figure}[t]
  \centering
  \includegraphics[trim=40 38 28 58, clip, width=\textwidth]{jss3813-plotsvlpred-1}
  \caption{Multi-step ahead predictive distributions (solid, gray and black) and observations (dashed, red).}
  \label{fig:svlpred}
\end{figure}

\subsection{Rolling window estimation}

Inspired by \fct{ugarchroll} in the \proglang{R} package \pkg{rugarch}~\citep{rugarch}, we introduce the suite of wrapper functions \fct{svsample\_roll}, \fct{svtsample\_roll}, \fct{svlsample\_roll}, and \fct{svtlsample\_roll}, built around their corresponding routines \fct{svsample}, \fct{svtsample}, \fct{svlsample}, and, respectively, \fct{svtlsample}, to simplify rolling window estimation of SV models.
In this estimation method, either a fixed width time window is \emph{moving} through the time series or a sequence of \emph{expanding} time windows with the same starting time point covers larger and larger chunks of the observations, and the same model is estimated in all time windows independently.
Next, each estimated model is employed for out-of-sample prediction, typically one day to one week ahead of the time window.
Lastly, the set of predicted values might be used to evaluate the model fit.

In Bayesian statistics, a natural approach for assessing the predictive power of a model is through its \emph{posterior predictive distribution}.
Its density, also called the \emph{predictive density}, is defined as
\begin{equation}\label{eq:predictive-density}
  p(y_{t+1}\mid\bm{y^o}_{[1:t]})=\int_{\bm{\mathcal{K}}}p(y_{t+1}\mid\bm{y^o}_{[1:t]}, \bm\kappa) p(\bm\kappa\mid\bm{y^o}_{[1:t]}) \text{d}\bm\kappa,
\end{equation}
where $\bm\kappa$ collects all unobserved variables, i.e., $\bm\kappa=(\mupar,\phipar,\sigmapar,\rhopar,\nupar,\hvec,\bm\beta)^\top$ in the most general case of SVtl, and the domain of integration $\bm{\mathcal{K}}$ is the set of all possible values for $\bm\kappa$.
We follow~\citet{geweke2010comparing} in our notation by using a superscript $\bm o$ for the vector of observed values $\bm{y^o}_{[1:t]}=(y_1,y_2,\dots,y_t)^\top$.
Equation~\ref{eq:predictive-density} can be seen as the integration of the predictive likelihood over the posterior distribution of all parameters and therefore it accounts for posterior parameter uncertainty for the predicted values.

The integral in Equation~\ref{eq:predictive-density} has no closed form and its dimensionality increases with $t$; it is intractable.
Hence, we rely on Monte Carlo integration and we simulate from the posterior predictive distribution.
For the evaluation of the predictive density at an observation $x=y^o_{t+1}$, called the \emph{predictive likelihood}, we apply the computation
\begin{equation}\label{eq:predictive-simulation}
  p(x\mid\bm{y^o}_{[1:t]}) \approx \frac1{M}\sum_{m=1}^M p(x\mid\bm{y^o}_{[1:t]},\bm{\kappa^{(m)}}),
\end{equation}
where $\bm{\kappa^{(m)}}$ denotes the $m$th posterior sample from the estimation procedure of the SV model.
For other applications, the quantiles of the posterior predictive distribution, henceforth the \emph{predictive quantiles}, might be of interest.
We estimate the $q\%$ quantile through random variates simulated from $y_{t+1}\sim p(y_{t+1}\mid\bm{y^o}_{[1:t]})$, which we acquire by repeating two steps for $m=1,\dots,M$:
\begin{itemize}
  \item[Step 1.] Simulate $\bm{\kappa^{(m)}}$ from the SV posterior $p(\bm\kappa\mid\bm{y^o}_{[1:t]})$, and
  \item[Step 2.] Simulate $y^{(m)}_{t+1}$ from $p(y_{t+1}\mid\bm{y^o}_{[1:t]}, \bm\kappa^{(m)})$.
\end{itemize}
Lastly, we take the $q\%$ quantile of the sample vector $(y^{(1)}_{t+1}, y^{(2)}_{t+1}, \dots, y^{(M)}_{t+1})^\top$ as the approximate $q\%$ quantile of the predictive density.
We implement the estimation of both the predictive likelihood and predictive quantiles in \stochvol{}.

All four rolling window routines \fct{svsample\_roll}, \fct{svtsample\_roll}, \fct{svlsample\_roll}, and \fct{svtlsample\_roll} bear the same programming interface.
They expect as their first argument the input data $\yvec_{[1:L]}$, which is of length $L$.
For estimating the SV model in each time window $j=1,\dots,J$ in the moving or expanding window scheme, the sub-vector $\yvec_{[j:(t+j-1)]}$, or, respectively, $\yvec_{[1:(t+j-1)]}$, is taken as data and is used to predict $n_\text{ahead}\ge1$ time steps ahead.
The width $t$ of the first time window can be determined from $L$, $J$, and $n_\text{ahead}$.
The following example demonstrates how the rolling window sampling routines can be called in \stochvol{}.
<<svroll, echo=TRUE, results='hide'>>=
set.seed(5)
res <- svsample_roll(CHF_logret, n_ahead = 1, forecast_length = 30,
  refit_window = "moving", calculate_quantile = c(0.01, 0.05),
  calculate_predictive_likelihood = TRUE)
@
Argument \code{n\_ahead} is used to set $n_\text{ahead}$, \code{forecast\_length} is used to set $J$, and \code{refit\_window} expects either \code{"moving"} or \code{"expanding"} to set the rolling window scheme to moving or expanding, respectively.
Argument \code{calculate\_quantile} expects a vector of numbers between 0 and 1; the numbers are interpreted as the quantiles to be predicted.
Furthermore, if \code{calculate\_predictive\_likelihood} is set to \code{TRUE}, the function estimates the predictive likelihood.
Lastly, the output \code{res} is a list of length $J$, i.e., one element for each time window.
It contains the respective posterior quantile and predictive likelihood results together with all posterior parameter draws for $\bm\kappa$.

\subsection{Specifying the prior hyperparameters}
As discussed in Section~\ref{sec:svpriors}, the prior distributions need to be specified before the estimation process can start.
Concerning the common model parameters $\mupar$, $\phipar$, and $\sigmapar$, all of \fct{svsample}, \fct{svtsample}, \fct{svlsample}, and \fct{svtlsample} expect through their input arguments \code{priormu}, \code{priorphi}, and \code{priorsigma} values for $(\mupriorf,\sqrt{\mupriors})$, $(\phipriorf,\phipriors)$, and $\sigmaprior$, respectively.
Furthermore, all sampling functions accept the argument \code{priorbeta} to set an independent prior for the regression coefficients by providing $(b_\beta,s_\beta)$, where $b_\beta$ and $s_\beta$ are the common mean and, respectively, the common standard deviation. For a general multivariate normal distribution, the \fct{specify\_priors} interface exists, which we detail later in this Section.
The prior for $\nupar$ can be influenced in \fct{svtsample} and \fct{svtlsample} by passing $\nuprior$ as the argument \code{priornu}.
Finally, \fct{svlsample} and \fct{svtlsample} take the \code{numeric} sequence $(\rhopriorf,\rhopriors)$ through the input argument \code{priorrho}.

The code snippet below shows all the default values of the prior hyperparameters.
<<printpriordefault, echo=TRUE, results='hide'>>=
svsample(CHF_logret, priormu = c(0, 100), priorphi = c(5, 1.5),
  priorsigma = 1, priorbeta = c(0, 10000))
svtsample(CHF_logret, priormu = c(0, 100), priorphi = c(5, 1.5),
  priorsigma = 1, priorbeta = c(0, 10000), priornu = 0.1)
svlsample(CHF_logret, priormu = c(0, 100), priorphi = c(5, 1.5),
  priorsigma = 1, priorbeta = c(0, 10000), priorrho = c(4, 4))
svtlsample(CHF_logret, priormu = c(0, 100), priorphi = c(5, 1.5),
  priorsigma = 1, priorbeta = c(0, 10000), priornu = 0.1,
  priorrho = c(4, 4))
@

As an alternative to the concise interface above, a broader set of prior distributions can be specified via an object created by the \fct{specify\_priors} function.
The function has an input argument for each model parameter: \code{mu}, \code{phi}, \code{sigma2}, \code{nu}, \code{rho}, \code{beta}, and additionally one for the variance of $\hpar_0$ called \code{latent0\_variance}.
There is a list of accompanying functions that create distributions in \stochvol{}:
\fct{sv\_beta} has arguments \code{shape1} and \code{shape2} and it is accepted for \code{phi} and \code{rho};
\fct{sv\_constant} has argument \code{value} and it is accepted for \code{mu}, \code{phi}, \code{sigma2}, \code{nu}, \code{rho}, and \code{latent0\_variance};
\fct{sv\_normal} has arguments \code{mean} and \code{sd} and it is accepted for \code{mu} and \code{phi};
\fct{sv\_multinormal} has arguments \code{mean} and either \code{sd} and \code{dim} or \code{precision}, and it is accepted for \code{beta};
\fct{sv\_exponential} has argument \code{rate} and it accepted for \code{nu};
\fct{sv\_gamma} has arguments \code{shape} and \code{rate} and it is accepted for \code{sigma2};
\fct{sv\_inverse\_gamma} has arguments \code{shape} and \code{scale} and it is accepted for \code{sigma2};
and \fct{sv\_infinity} has no arguments and it is accepted for \code{nu} hence turning the Student's $t$~distribution into a normal distribution.
Additionally, \code{latent0\_variance} accepts the character value \code{"stationary"}.
All four sampling methods accept the prior specification object through the input argument \code{priorspec}.

All input arguments for \code{specify\_priors} are optional, their default values and how they are used is seen below.
<<printpriorspecdefault, echo=TRUE, results='hide'>>=
ps <- specify_priors(mu = sv_normal(mean = 0, sd = 100),
  phi = sv_beta(shape1 = 5, shape2 = 1.5), rho = sv_constant(0),
  sigma2 = sv_gamma(shape = 0.5, rate = 0.5), nu = sv_infinity(),
  beta = sv_multinormal(mean = 0, sd = 10000, dim = 1),
  latent0_variance = "stationary")
svsample(CHF_logret, priorspec = ps)
@

\subsection{Setting up the Markov chain}

When conducting Bayesian inference using an MCMC sampling scheme, the number of draws from the posterior distribution, the length of the so-called burn-in phase, the initial values of the Markov chain, and the various strategies of storing the results are all of general interest.
The input arguments \code{draws} and \code{burnin} settle the first two points.
A sample size of \code{burnin + draws} is acquired from the posterior distribution out of which the first \code{burnin} number of draws are thrown away.
The default is to draw 10000 elements after a burnin of 1000 for SV models without leverage, and draw 20000 elements after a burnin of 2000 for SV models with leverage, which in our experience is enough for most applications.

As for the initial values, \code{startpara} and \code{startlatent} provide a way to set them.
The argument \code{startpara} is expected to be a named \code{list} mapping parameter names to starting values, and \code{startlatent} must be a sequence of length $\dimy$ that contains starting values for $\hvec$.
Default values are set to be the prior mean for $\phipar$, $\sigmapar$, $\nupar$, and $\rhopar$, these have only minor influence on the Markov chain.
The default value for $\bm\beta$ is the ordinary least squares estimator $\bm{(X^\top X)^{-1}\bm{X^\top}}\yvec$, where $\bm X$ denotes the regression design matrix and $\yvec$ denotes the vector of observations.
After setting $\bm\beta$, the level of log-variance $\mupar$ is initialized according to the Bayesian linear regression
\begin{equation}\label{eq:initmu}
  \begin{split}
    \log(y_t^2) &= \mupar+\xi_t, \\
    \mupar &\sim \Normal{\mupriorf}{\mupriors},
  \end{split}
\end{equation}
where $\xi_t\sim\Normal{-1.27}{4.934}$.
Equation~\ref{eq:initmu} results from the first line of Equation~\ref{eq:vanillasv} by fixing $h_t$ at its stationary expected value $\mupar$ and then taking $x\mapsto\log(x^2)$ of both sides.
The homoskedastic error term $\xi_t$ is acquired as the Laplace approximation to $\log(\varepsilon_t^2)$~\citep{harvey1996estimation}.
At the end, by default all values of the vector \code{startlatent} are set to the initial value of $\mupar$.

It is customary to start independent Markov chains in parallel and \stochvol{} provides facilities for that in all of its sampling procedures.
The argument \code{n\_chains} is expected to be a positive integer, it sets the number of independent chains.
Additionally, arguments \code{parallel}, \code{n\_cpus}, and \code{cl} can be used to control parallelism used by \stochvol{}.
To overwrite the default sequential execution strategy, \code{parallel} is to be set either to \code{"snow"}, to employ the so-called ``SNOW'' clusters, or to \code{"multicore"} to use the ``multicore'' type computation~\citep{rlanguage}.
Next, argument \code{n\_cpus} should be set to the physical number of parallel processing units to be used.
Finally, in case ``SNOW'' is applied, the sampling routines optionally accept an already running ``SNOW'' cluster through argument \code{cl}.

As mentioned earlier, the sampling algorithms for the latent states $\hvec$ in \stochvol{} rely on a Gaussian mixture approximation as in~\citet{omori2007stochastic} and~\citet{kastner2014ancillarity}.
The approximation tends to be very good, therefore the default setting is not to correct for model misspecification.
However, this correction can be enabled in all of the sampling routines through the \code{expert} argument as shown for \fct{svsample} in the following.
<<results='hide'>>=
y <- svsim(50)$y
svsample(y, expert = list(correct_model_misspecification = TRUE))
@

Lastly, \stochvol{} provides three ways to economize storage during and after the execution of the sampler.
Setting the \code{integer} argument \code{thinpara} to $\iota$ tells the sampler to store only every $\iota$th draw of the vector of parameters, and supplying a value for \code{thinlatent} does the same for $\hvec$.
Finally, one has the opportunity not to store the full vector $\hvec$ but only its last value by setting \code{keeptime = "last"}.
The default behavior is to store every draw after the burn-in phase.

\section[The factorstochvol package]{The \factorstochvol{} package} \label{sec:factorstochvol}

The most common workflow of using \factorstochvol{} for fitting multivariate factor SV models consists of the following steps:
(1) Prepare the data,
(2) decide on an identification structure,
(3) specify the prior hyperparameters,
(4) run the sampler,
(5) investigate the output and visualize the results, and
(6) predict (if required).
These steps are described in detail in the following sections.

\subsection{Preparing the data} \label{sec:fsvuse1}

The workhorse in \factorstochvol{} is the sampling function \fct{fsvsample}. It expects the data to come in form of a matrix $\bm Y=(\yvec_1,\dots,\yvec_\leny)^\top$ with $\leny$ rows and $\dimy$ columns.
For illustration, we use the exchange rate data set in \stochvol{} which contains $3140$ daily observations of exchanges rates for $23$ currency pairs against EUR, ranging from March 3, 2000 to April 4, 2012.
To keep the analysis simple and computation times moderate, we however only model the last $1001$ days of the first six series in alphabetical order (Australian dollar, Canadian dollar, Swiss franc, Czech koruna, Danish krone, Great British pound) for further analysis.
Instead of using the nominal exchange rates we compute log returns. This leaves us with a data set of size $n = 1000$ and $m = 6$.
The data is prepared using the code snippet below and visualized in Figure~\ref{fig:exrates} using the \pkg{zoo} package \citep{zoo}.
<<fsvprepdata, echo=TRUE, fig.width=10, fig.height=5, results='hide'>>=
library("factorstochvol")
library("zoo")
data("exrates", package = "stochvol")
m <- 6
n <- 1000
y <- 100 * logret(tail(exrates[, seq_len(m)], n + 1))
y <- zoo(y, order.by = tail(exrates$date, n))
plot(y, main = "", xlab = "Time")
@
\begin{figure}[t]
  \centering
  \includegraphics[trim=5 40 20 40, clip, width=\textwidth]{jss3813-fsvprepdata-1}
  \caption{Percentage log returns of six EUR exchange rates.}
  \label{fig:exrates}
\end{figure}

\subsection{Deciding on an identification structure}\label{sec:fsvuse2}

The likelihood in factor models is invariant to certain factor transformations such as reordering of factors and their loadings or sign switches thereof. In addition to this, it is often multimodal. Consequently, identifying the factor loadings is far from  trivial.
The most common way to address this issue in factor SV models is to impose a lower-diagonal factor loadings matrix where all elements above the diagonal are set to zero \citep[e.g.,][]{aguilar2000bayesian, chib2006analysis, han2006asset, zhou2014bayesian}.
To use this constraint in \factorstochvol, the argument \code{restrict = "upper"} can be passed to the main sampling function \fct{fsvsample}.
Evidently, this practice imposes an order dependence, as, e.g., the first variable is not allowed to load on anything else but the first factor.

A rather ad hoc method for automatically ordering the data is implemented in the helper function \fct{preorder}.
After a maximum likelihood factor model fit to the data (using \fct{factanal} from the \pkg{stats} package with the default \code{varimax} rotation), the series are ordered as follows: The variable with the highest loading on factor 1 is placed first, the variable with the highest loading on factor 2 second (unless this variable is already placed first, in which case the variable with the second highest loading is taken), et cetera.
For the data set at hand, this would imply the following ordering for a two-factor model.
<<preorder, echo=TRUE>>=
preorder(y, factors = 2)
@
According to this algorithm, the second series should be placed first and the third series should be placed second. Thereafter, the alphabetical ordering remains.

To achieve this effect without reordering the data, a \code{logical} matrix of size $\dimy \times r$ can be passed to \fct{fsvsample} via \code{restrict}, where the entry \code{TRUE} means that this element is restricted to zero; \code{FALSE} means that it is to be estimated from the data.
Similarly to \fct{preorder}, the function \fct{findrestrict} tries to automate this procedure.
Again, the maximum likelihood estimates from a static factor analysis are used; however, \fct{findrestrict} uses a slightly different algorithm than the one above:
The variable with the lowest absolute loadings on factors $2, 3, \dots, r$ (relative to factor 1) is determined to lead the first factor, the variable with the lowest absolute loadings on factors $3, 4, \dots, r$ (relative to factors 1 and 2) is placed second, et cetera.
Below is the result for the data set at hand.
<<findrestrict, echo=TRUE>>=
findrestrict(y, factors = 2)
@
If \fct{fsvsample} is called with the argument \code{restrict = "auto"}, it automatically invokes \fct{findrestrict} with the appropriate number of factors.
Using \code{restrict = "none"} (the default) causes the sampler not to place any constraints on the loadings matrix; thus, the resulting posterior draws may be unstable or suffer from multiple local modes.
If, however, inference on the factor loadings themselves is not the primary concern of the analysis, leaving the factor loadings unidentified may be the preferred option.
This is in particular the case when inference for the covariance matrix is sought, as this only depends on $\Loadings$ through the rotation-invariant transformation of Equation~\ref{eq:vardecomp}.
For a more elaborate discussion of these issues, we refer the reader to \citet{sentana2001identification} who discuss automatic identification through hetero\-skedasticity. A comparison of log predictive scores under different identification schemes for factor SV models is given in \citet{kastner2017efficient}; see also \citet{fruhwirth2018sparse} for related issues in static factor models.
To continue with the current example, we chose not to place any a priori restrictions on the factor loadings matrix while using a row-wise normal-gamma shrinkage prior on the factor loadings matrix \citep[cf.][]{kastner2019sparse}.

\subsection{Specifying prior hyperparameters} \label{sec:fsvuse3}

Apart from the obvious prior choice about the number of factors and the identification scheme discussed above, a number of hyperparameter choices are available in \factorstochvol.
Regarding the log-variance processes, the interface is analogous to that of \fct{svsample}.
In the following, $i=1,\dots,\dimy$ and $j=1,\dots,\nfac$ index the idiosyncratic and the factor log-variance processes, respectively.
The pair of common prior hyperparameters $(b_\beta, B_\beta)$ can be passed as a sequence of length two to \code{priorbeta}.
The common prior of $\muidi_i$ can be set by passing a sequence of length two -- the mean and the standard deviation of the normal distribution -- to \code{priormu};
the common priors of $\phiidi_i$ and $\phifac_j$ can also be set by passing sequences of length two -- the parameters of the corresponding beta distribution -- to \code{priorphiidi} and to \code{priorphifac}, respectively;
similarly, the common priors of $\sigmaidi_i$ and $\sigmafac_j$ can be specified via the arguments \code{priorsigmaidi} and \code{priorsigmafac}, respectively, that accept as positive numbers the scale $\sigmaprior$ of the corresponding gamma distribution.

As discussed in Section~\ref{sec:fsvpriors}, \factorstochvol{} offers three specifications as priors for $\Loadings$, controlled through the argument \code{priorfacloadtype}.
To use the first option (\code{priorfacloadtype = "normal"}), one needs to fix the values of $\tau_{ij}$ a priori.
The user can pass these fixed values to \fct{fsvsample} via the argument \code{priorfacload}, either as an $\dimy\times\nfac$ matrix with positive entries or as a single positive number which will be recycled accordingly.
For the second option, the normal gamma prior with row-wise or column-wise shrinkage (\code{priorfacloadtype = "rowwiseng"} and \code{priorfacloadtype = "colwiseng"}, respectively), the value of argument \code{priorfacload} is then interpreted as the shrinkage parameter $a$.
Both specifications of the normal gamma prior need the values $c$ and $d$.
They can be set as a two-element vector passed to the argument \code{priorng}.

\subsection{Running the MCMC sampler}

Running the sampler corresponds to invoking \fct{fsvsample}. Apart from the prior settings discussed above, its most important arguments are listed below with the default value in brackets. For a complete list of all arguments and more details, see \code{?fsvsample}.
\begin{itemize}
  \item \code{y}: the data;
  \item \code{factors} [\code{1}]: the number of factors;
  \item \code{draws} [\code{1000}]: the number of MCMC samples to be drawn after burnin;
  \item \code{thin} [\code{1}]: the amount of thinning (every \code{thin}\textsuperscript{th} draw is kept);
  \item \code{burnin} [\code{1000}]: the length of the burn-in period, i.e., the number of MCMC draws to be discarded before the samples are considered to emerge from the stationary distribution,
  \item \code{zeromean} [\code{TRUE}]: a logical value indicating whether $\bm\beta$ is to be estimated from the data or whether $\bm\beta$ is set to zero (the default);
  \item \code{keeptime} [\code{"last"}]: either \code{"all"}, meaning that all latent log volatilities are being monitored at all points in time, or \code{"last"}, meaning that the latent log volatility draws are only stored at $t = n$, the last point in time; the latter setting is the default to avoid excessive memory usage in higher dimensions;
  \item \code{heteroskedastic} [\code{TRUE}]: indicator(s) to turn off stochastic volatility for the idiosyncratic variances, the factor variances, or both;
  \item \code{samplefac} [\code{TRUE}]: indicator to turn off sampling of the factors; useful to work with observed instead of latent factors \citep[see][for a use case of this]{kastner2019sparse};
  \item \code{runningstore} [\code{6}]: to avoid having to store all MCMC draws, \code{fsvsample}'s default is to compute and store the first two ergodic moments of some interesting quantities (namely log variances, factors, volatilities, covariance matrices, correlation matrices, communalities) only; the default (\code{runningstore = 6}) is to compute and store everything; however, one can set \code{runningstore} to a lower number to save computation time; the argument \code{runningstoremoments} [\code{2}] can further be used to modify the number of moments to be stored;
  \item \code{runningstorethin} [\code{10}]: indicates how often ergodic moments should be calculated, where \code{1} means that this should be done at every iteration and higher numbers lessen both runtime as well as accuracy;
  \item \code{quiet} [\code{FALSE}]: a logical indicator determining the verbosity of \code{fsvsample}.
\end{itemize}

For our illustrative example, most settings are left at their default values. The number of factors is increased from one to two, instead of 1000 we sample 10000 draws, we estimate a constant mean, a thinning of 10 is used, and \code{quiet} is set to \code{TRUE}.
<<runmodel, echo=TRUE>>=
set.seed(1)
res <- fsvsample(y, factors = 2, draws = 10000, zeromean = FALSE,
  thin = 10, quiet = TRUE)
@

\subsection{Investigating the output and visualizing the results}

The resulting object
<<printrres, echo = TRUE>>=
res
@
holds a rich amount of information. In particular, it contains
\begin{itemize}
  \item draws of certain posterior quantities such as the factors $\fvec$, the factor loadings $\Loadings$, the various factor and idiosyncratic SV parameters, the latent factor and idiosyncratic log variances $\bm{\hfac}$ and $\bm{\hidi}$, and the intercept $\bm\beta$,
  \item configuration settings such as the number of draws, potential restrictions on the loadings matrix, prior hyperparameters, etc.,
  \item running moments (such as means and standard deviations) of quantities of interest, depending on the values of \code{runningstore} and \code{runningstoremoments} specified when calling \fct{fsvsample},
  \item the data input $\bm{y}$.
\end{itemize}
For more details, please investigate \code{str(res)} and/or \code{?fsvsample}.

Using \fct{covmat}, one can extract the MCMC draws of the implied covariance matrices for all points in time which have been stored during sampling. By default, this is the last point in time (\code{keeptime = "last"}), and thus
<<covn, echo = TRUE>>=
dim(cov_n <- covmat(res))
@
shows that we have stored $\Sexpr{res$config$draws/res$config$thin}$ posterior draws of a $\Sexpr{m}\times\Sexpr{m}$ covariance matrix at one point in time, $t = n = \Sexpr{n}$. To check convergence, one can take a look at the trace plot and the autocorrelation function of the log determinant, i.e.,
<<logdetcovn, echo = 2:5, fig.width = 10, fig.height=3.5, results='hide'>>=
par(mfrow = c(1, 2), mgp = c(1.7, 0.5, 0), mar = c(3, 3, 1, 1))
logdet <- function (x) log(det(x))
logdet_n <- apply(cov_n[,,,1], 3, logdet)
ts.plot(logdet_n)
acf(logdet_n, main = "")
@
The results are visualized in Figure~\ref{fig:logdetcovn}; decent mixing for this quantity is apparent.

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth, clip, trim=5 5 10 10]{jss3813-logdetcovn-1}
  \caption{Trace plot and empirical autocorrelation function of the log determinant of the model-implied covariance matrix at $t = n$.}
  \label{fig:logdetcovn}
\end{figure}

To assess the mixing speed of each individual covariance matrix element, one can check, e.g., the estimated effective sample size (out of $\Sexpr{res$config$draws/res$config$thin}$ draws kept) which is implemented in \pkg{coda}. Again, no major convergence problems are apparent.
<<covess, echo = TRUE>>=
round(apply(cov_n, 1:2, coda::effectiveSize))
@

Assuming that \code{runningstore} was set sufficiently high when sampling, several convenience functions can be used for quick visualizations without having to post-process the MCMC draws. For example, to visualize the time-varying correlation matrices, consider
<<corimageplot, echo=2, results='hide'>>=
par(mfrow = c(1, 3), xpd = TRUE)
corimageplot(res, these = seq(1, n, length.out = 3), plotCI = "circle",
  plotdatedist = 2, date.cex = 1.1)
@
\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth, clip, trim=22 185 0 174]{jss3813-corimageplot-1}
  \caption{Three estimated correlation matrices and their posterior uncertainty depicted using circles.
    Inner (outer) radii of the circles illustrate to the posterior mean minus (plus) two standard deviations.
    Colors blue and red represent negative and positive values, respectively.
    Furthermore, the transparency of the circles represents the posterior mean.
  Note that the diagonal is left white as it trivially contains ones (without uncertainty).}
  \label{fig:corimageplot}
\end{figure}
which produces the three estimated posterior correlation matrices depicted in Figure~\ref{fig:corimageplot}. Setting \code{plotCI = "circle"} visualizes posterior uncertainty -- inner and outer radii correspond to the posterior mean plus/minus two standard deviations, respectively.

To get an idea about how the marginal volatilities evolve over time, \fct{voltimeplot} can be used. To exemplify,
<<voltimeplot, echo=2:3, fig.width = 10, fig.height = 3, cache.rebuild = TRUE, results='hide'>>=
par(mgp = c(1.7, 0.5, 0), mar = c(2, 1.5, 1, 0.5))
palette(RColorBrewer::brewer.pal(7, "Dark2")[-5])
voltimeplot(res, legend = "top")
@
\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth, clip, trim=0 10 2 10]{jss3813-voltimeplot-1}
  \caption{Posterior means of daily marginal volatilities in percent.}
  \label{fig:voltimeplot}
\end{figure}
yields the estimated volatilities in Figure~\ref{fig:voltimeplot}. The financial crisis of 2008 and the capping of CHF's appreciation in September 2011 are clearly visible, while DKK's volatility (relative to EUR) is practically zero. Note that \fct{voltimeplot} respects palette changes. In the above example, \pkg{RcolorBrewer} \citep{rcolorbrewer} is used. Moreover,
<<cortimeplot, echo=2:4, fig.width = 10, fig.height = 5, cache.rebuild = TRUE, results='hide'>>=
par(mfrow = c(2, 1), mgp = c(1.7, 0.5, 0), mar = c(2, 1.5, 1, 0.5))
palette(RColorBrewer::brewer.pal(6, "Dark2"))
cortimeplot(res, 1)
cortimeplot(res, 2)
@
\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth, clip, trim=4 4 0 4]{jss3813-cortimeplot-1}
  \caption{Posterior means of correlations with AUD (top panel) and CAD (bottom panel).}
  \label{fig:cortimeplot}
\end{figure}
yields the estimated pairwise correlations in Figure~\ref{fig:cortimeplot}. While, relative to EUR, the estimated correlation between AUD and CAD appears to be relatively stable over time, correlations with CHF can become negative at times. To visualize the \emph{communalities}, i.e., the proportions of variances explained through the latent factors, invoke
<<comtimeplot, echo=2, fig.height = 6.5, results='hide'>>=
par(mgp = c(1.7, 0.5, 0), mar = c(3, 3, 1, 1))
comtimeplot(res, maxrows = 6)
@
\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{jss3813-comtimeplot-1}
  \caption{Communalities: Posterior means plus/minus two posterior standard deviations.
    The six panels correspond to the six observation series and they depict the percentage of volatility explained by the latent factors.
    The idiosyncratic volatility attributes for the unexplained part.}
  \label{fig:comtimeplot}
\end{figure}
which yields the estimated communalities in Figure~\ref{fig:comtimeplot}.

To gain an even deeper understanding of the estimated model, we now turn towards examining the latent factors and their variances themselves.
To visualize the loadings, the functions \fct{facloadpairplot}, \fct{facloadcredplot}, \fct{facloadpointplot}, \fct{facloadtraceplot}, and \fct{facloaddensplot} are available; the former two are exemplified in Figure~\ref{fig:loadplot2}.
Moreover, we can see the factor log variances produced through \code{logvartimeplot(res, show = "fac")}.
Similarly, \code{logvartimeplot(res, show = "idi")} produces plots of the idiosyncratic log variances which are displayed in Figure~\ref{fig:idivarplot}.
<<loadplot2, echo=2:3, fig.width=4.5, fig.height=4.5, results='hide'>>=
par(mgp = c(1.7, 0.5, 0), mar = c(2.7, 2.7, 2, 0.5))
facloadpairplot(res)
facloadcredplot(res)
@
<<varplot, echo=2, fig.width=10, fig.height=4, results='hide'>>=
par(mgp = c(1.7, 0.5, 0), mar = c(2.7, 2.7, 2, 0.5))
logvartimeplot(res, show = "fac")
@
<<varplot2, echo=2, fig.width=7, fig.height=6.5, results='hide'>>=
par(mgp = c(1.7, 0.5, 0), mar = c(2.7, 2.7, 2, 0.5))
logvartimeplot(res, show = "idi", maxrows = 6)
@
\begin{figure}[tp]
  \centering
  \includegraphics[width=0.49\textwidth]{jss3813-loadplot2-1}\hfill
  \includegraphics[width=0.49\textwidth]{jss3813-loadplot2-2}\\
  \includegraphics[width=\textwidth]{jss3813-varplot-1}
  \caption{Factor loadings and factor variances.
    The top panels visualize the joint posterior distribution of the two factor loadings.
    While the colored clouds on the left consist of posterior draws and provide details, the posterior $0.01$/$0.99$ credible intervals on the right summarize the marginal distributions.
    The bottom panels depict posterior means plus/minus two standard deviations of factor log variances.}
  \label{fig:loadplot2}
\end{figure}
\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{jss3813-varplot2-1}
  \caption{Idiosyncratic log variances: Posterior means plus/minus two standard deviations.
    Panels illustrate the marginal posterior distributions of the diagonal elements of $\Sigmaidi_t$ in Equation~\ref{eq:fsvcov}.}
  \label{fig:idivarplot}
\end{figure}

In addition to the above, there is the plotting function \fct{paratraceplot} which produces trace plots of all parameters associated with the log variances processes: mean, persistence, and volatility of log variances.

In order to provide some guidance when it comes to selecting the number of factors, \factorstochvol{} ships with the function \fct{evdiag}.
It computes and visualizes the eigenvalues of $\Loadings^\top\Loadings$ which can be used as a rough guide in analogy to a scree plot for static factor models.
To use it, one can fit a model with a relatively large number of factors and assess the importance of each of these (in descending order).
For example, the code below produces Figure~\ref{fig:evdiag}, hinting at a model with no more than four factors.

<<evdiag, fig.width=10, fig.height=4, echo=2:4, results = 'hide'>>=
par(mgp = c(1.7, 0.5, 0), mar = c(2.7, 2.7, 2, 0.5))
set.seed(6)
largemodel <- fsvsample(y, factors = 6)
evdiag(largemodel)
@
\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{jss3813-evdiag-1}
  \caption{Eigenvalues of $\Loadings^\top\Loadings$ which can be used as a rough guide to selecting the number of factors.}
  \label{fig:evdiag}
\end{figure}

\subsection{Predicting covariances, correlations, and future observations}

One of the main use cases of \factorstochvol{} might be to predict covariance and correlation matrices of time series. To this end, \fct{predcov} and \fct{predcor} yield draws from the posterior predictive distribution of these, defined in analogy to Equation~\ref{eq:predictive-density}. For instance, the code below can be used to obtain one-step-ahead posterior predictive means and standard deviations for the correlation matrix on April 5, 2012 (using data up to April 4 only).
<<predcov1, echo=TRUE>>=
set.seed(4)
predcor1 <- predcor(res)
round(apply(predcor1[,,,1], 1:2, mean), 2)
round(apply(predcor1[,,,1], 1:2, sd), 2)
@

To obtain draws from the posterior predictive distribution of new data points, one can simply draw from the corresponding mixture of multivariate normals. In Figure~\ref{fig:preddist}, these draws are visualized via \fct{heatpairs} from \pkg{LSD} \citep{LSD}.
<<preddist, fig.height = 6, fig.width = 9, echo = TRUE, results='hide'>>=
set.seed(5)
predcov_1 <- predcov(res)
preddraws <- t(res$beta)
for (i in seq_len(NROW(preddraws)))
  preddraws[i,] <- preddraws[i,] + chol(predcov_1[,,i,1]) %*% rnorm(m)
plotlims <- quantile(preddraws, c(0.01, 0.99))
LSD::heatpairs(preddraws, labels = colnames(y), cor.cex = 1.5, gap = 0.3,
  xlim = plotlims, ylim = plotlims)
@
\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth, trim=20 20 20 40, clip]{jss3813-preddist-1}
  \caption{Predictive draws from the one-step-ahead posterior predictive distribution (above the diagonal) and empirical correlation coefficients (below the diagonal).}
  \label{fig:preddist}
\end{figure}

To conclude, we note that convenience functions such as \fct{predloglik} or \fct{predloglikWB} may aid in approximating predictive likelihoods \citep[cf., e.g.,][]{geweke2010comparing}. For instance, assuming that the actually observed value of $\bm y_{n+1} = \bm y_{n+2} = (0,0,0,0,0,0)^\top$, we can approximate the one and two step ahead log predictive scores through code along the following lines.
<<echo = TRUE>>=
set.seed(6)
predloglik(res, matrix(0, nrow = 2, ncol = m), ahead = 1:2, each = 10)
@
For a more detailed analysis of the out-of-sample performance of \factorstochvol{} for exchange rate data, we refer to \cite{kastner2017efficient}; for a predictive exercise on stock market data, please see \cite{kastner2019sparse}.

\section{Summary and discussion} \label{sec:summary}

We extended the work of~\citet{kastner2016dealing} to other SV models, including the univariate heavy-tailed SV, the SV model with leverage, and the multivariate factor SV model.
We showcased the features that are the most important to end users in \proglang{R}: estimation through the sampler functions, visualization, summary, and prediction methods.
Due to its more involved nature, however, we did not include the description of the \proglang{C++} interface.
Two functions called \fct{update\_fast\_sv} and \fct{update\_general\_sv} are exported and programmers have the possibility to access the samplers in \stochvol{} directly from \proglang{C++} after linking to the compiled package.
For usage examples, see the implementations of \factorstochvol{} or \pkg{shrinkTVP} \citep{shrinktvp}.
Source code is available in the GitHub repositories \url{https://github.com/gregorkastner/stochvol} and \url{https://github.com/gregorkastner/factorstochvol}.

\section*{Acknowledgments}

Gregor Kastner acknowledges funding from the Austrian Science Fund (FWF) for the project ``High-dimensional statistical learning: New methods to advance economic and sustainability policy'' (ZK 35), jointly carried out by the University of Klagenfurt, WU Vienna University of Economics and Business, Paris Lodron University Salzburg, TU Wien, and the Austrian Institute of Economic Research (WIFO).

\bibliography{ref2}

\end{document}

### 2021-05-07 R/exports.R

@@ -2,35 +2,35 @@
 #  R package stochvol by
 #     Gregor Kastner Copyright (C) 2013-2020
 #     Darjus Hosszejni Copyright (C) 2019-2020
-#  
+#
 #  This file is part of the R package stochvol: Efficient Bayesian
 #  Inference for Stochastic Volatility Models.
-#  
+#
 #  The R package stochvol is free software: you can redistribute it
 #  and/or modify it under the terms of the GNU General Public License
 #  as published by the Free Software Foundation, either version 2 or
 #  any later version of the License.
-#  
+#
 #  The R package stochvol is distributed in the hope that it will be
 #  useful, but WITHOUT ANY WARRANTY; without even the implied warranty
 #  of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the GNU
 #  General Public License for more details.
-#  
+#
 #  You should have received a copy of the GNU General Public License
 #  along with the R package stochvol. If that is not the case, please
 #  refer to <http://www.gnu.org/licenses/>.
 #  #####################################################################################
 
 #' Bindings to \code{C++} Functions in \code{stochvol}
-#' 
+#'
 #' All the heavy lifting in \code{stochvol} is implemented in \code{C++}
 #' with the help of \code{R} packages \code{Rcpp} and \code{RcppArmadillo}.
 #' These functions call the MCMC samplers in \code{C++} directly without any
 #' any validation and transformations, expert use only!
-#' 
+#'
 #' The sampling functions are separated into fast SV and general SV. See more details
 #' in the sections below.
-#' 
+#'
 #' @param y numeric vector of the observations
 #' @param draws single positive integer, the number of draws to
 #' return (after the burn-in)
@@ -101,7 +101,7 @@
 #' efficiency of the model parameters, and it employs all-without-a-loop (AWOL)
 #' for computationally efficient Kalman filtering of the conditionally Gaussian state space.
 #' Correction for model misspecification happens as a post-processing step.
-#' 
+#'
 #' Fast SV employs sampling strategies that have been fine-tuned and specified for
 #' vanilla SV (no leverage), and hence it can be fast and efficient but also more limited
[…]

### 2021-05-07 article2.Rnw
update article; remove some dependencies; regenerate manual

@@
   useFancyQuotes = FALSE)
+
+library("RColorBrewer")
+library("LSD")
+library("zoo")
 @
@@
 <<results='hide'>>=
+set.seed(3)
 y <- svsim(50)$y
@@
 
-\bibliography{ref2}
+\bibliography{ref}
 

### 2021-05-07 article.Rnw
both vignettes compile and are transferred to the package during building

@@
 %\VignetteIndexEntry{Dealing with Stochastic Volatility in Time Series Using the R Package stochvol}
-
 %\VignetteKeyword{Bayesian inference}
@@
 %\VignetteKeyword{exchange rates}
+%\VignetteEngine{knitr::knitr}
+%\VignetteEncoding{UTF-8}
 
@@
 This vignette corresponds to an article of the same name which is published in the Journal of Statistical Software.
-The code base of the vignette has been updated to reflect the changes in recent versions of \stochvol{}.
+The code base of the vignette has been updated to reflect the changes in recent versions of \pkg{stochvol}.
 To cite, please use \cite{kas:dea}.
@@
 \begin{center}
- \includegraphics[width=\textwidth]{Figures/jss2014-usd1}
+ \includegraphics[width=\textwidth]{Figures/jss2014-usd1-1}
  \caption{Visualization of EUR-USD exchange rates included in the \pkg{stochvol} package.}
@@
 \begin{center}
- \includegraphics[width=\textwidth]{Figures/jss2014-usd2}
+ \includegraphics[width=\textwidth]{Figures/jss2014-usd2-1}
  \caption{Visualization of a simulated time series as provided by the default \code{plot} method.}
@@
 \begin{center}
- \includegraphics[width=\textwidth]{Figures/jss2014-usd3}
+ \includegraphics[width=\textwidth]{Figures/jss2014-usd3-1}
  \caption{Visualization of estimated contemporaneous volatilities of EUR-USD exchange rates, as provided by \code{volplot}.
@@
 \begin{center}
- \includegraphics[width=\textwidth]{Figures/jss2014-usd4}
+ \includegraphics[width=\textwidth]{Figures/jss2014-usd4-1}
  \caption{As above, now with medians (black line) and 1\%/10\%/90\%/99\% quantiles (gray lines). This behavior can be achieved through a preceding call of \code{updatesummary}.}
@@
 \begin{center}
- \includegraphics[width=\textwidth]{Figures/jss2014-usd5}
+ \includegraphics[width=\textwidth]{Figures/jss2014-usd5-1}
  \caption{Trace plots of posterior draws for the parameters $\mupar, \phipar, \sigmapar$.}
@@
 \begin{center}
- \includegraphics[width=\textwidth]{Figures/jss2014-usd6}
+ \includegraphics[width=\textwidth]{Figures/jss2014-usd6-1}
  \caption{Posterior density estimates (black solid lines) along with prior densities (dashed gray lines). Individual posterior draws are indicated by the underlying rug.}
@@
 \begin{center}
- \includegraphics[width=\textwidth]{Figures/jss2014-usd7}
+ \includegraphics[width=\textwidth]{Figures/jss2014-usd7-1}
  \caption{Illustration of the default \code{plot} method for \code{svdraws}-objects.
@@
 \begin{center}
- \includegraphics[width=\textwidth]{Figures/jss2014-usd8}
+ \includegraphics[width=\textwidth]{Figures/jss2014-usd8-1}
  \caption{Mean standardized residual plots for assessing the model fit, as provided by the corresponding \code{plot} method.
@@
 \begin{center}
-\includegraphics[width=\textwidth]{Figures/jss2014-homo}
+\includegraphics[width=\textwidth]{Figures/jss2014-homo-1}
 \caption{Trace plots and kernel density estimates for some draws from the marginal posterior distributions in the regression model with heteroskedastic errors.
@@
 Instead of homoskedastic errors, we now specify the error covariance matrix in Equation~\ref{reg} to be $\Sigmab \equiv \diag{\e^{h_1}, \dots, \e^{h_n}}$, thus introducing nonlinear dependence between the observations due to the AR(1)-nature of $\hb$.
-Instead of cooking up an entire new sampler, we adapt the code from above utilizing the \pkg{stochvol} package.\footnote{Since \stochvol{} version 3.0, regression is a built-in feature of the function \code{svsample_fast_cpp} through its argument \code{designmatrix}.} To do so, we simply replace the sampling step of $\sigmaepsextrasuper{2}$ from an Inverse-Gamma distribution with a sampling step of $\allpara$ and $\hb$ through a call to \code{svsample_fast_cpp}.\footnote{In earlier version of the \pkg{stochvol} package, this function was called \code{svsample2}.} This function is a minimal-overhead version of the regular \code{svsample}.
+Instead of cooking up an entire new sampler, we adapt the code from above utilizing the \pkg{stochvol} package.\footnote{Since \pkg{stochvol} version 3.0, regression is a built-in feature of the function \code{svsample\_fast\_cpp} through its argument
+\code{designmatrix}.} To do so, we simply replace the sampling step of $\sigmaepsextrasuper{2}$ from an Inverse-Gamma distribution with a sampling step of $\allpara$ and $\hb$ through a call to \code{svsample\_fast\_cpp}.\footnote{In earlier version of the \pkg{stochvol} package, this function was called \code{svsample2}.} This function is a minimal-overhead version of the regular \code{svsample}.
 It provides the full sampling functionality of the original version but has slightly different default values, a simplified return value structure, and it does not perform costly input checks.
@@
   \begin{itemize}
-   \item the latent volatilities/parameters by conditioning on the regression parameters and calling \code{svsample2},
+   \item the latent volatilities/parameters by conditioning on the regression parameters and calling \code{svsample\_fast\_cpp},
    \item the regression parameters by conditioning on the latent volatilities and calling \code{rmvnorm}:
@@
  \begin{center}
-\includegraphics[width=\textwidth]{Figures/jss2014-hetero}
+\includegraphics[width=\textwidth]{Figures/jss2014-hetero-1}
 \caption{Trace plots and kernel density estimates
@@
 
-It should be noted that even though \code{svsample2} is considerably faster than \code{svsample}, the cost of interpreting this function in each MCMC iteration is still rather high (which lies in the nature of \proglang{R} as an interpreted language).
+It should be noted that even though \code{svsample\_fast\_cpp} is considerably faster than \code{svsample}, the cost of interpreting this function in each MCMC iteration is still rather high (which lies in the nature of \proglang{R} as an interpreted language).
 Thus, as of package version \code{0.8-0}, a single-step \code{update} is also made available at the \proglang{C}/\proglang{C++} level for samplers coded there.
@@
 \begin{center}
- \includegraphics[width=\textwidth]{Figures/jss2014-scatter}
+ \includegraphics[width=\textwidth]{Figures/jss2014-scatter-1}
 \end{center}
@@
 
-<<>>=
+<<results='hide'>>=
 if (usePreCalcResults) {
@@
 #configuration parameters
-draws <- 10000  #100000  TODO
+draws <- 100000
 thinning <- 100
@@
 \begin{center}
- \includegraphics[width=\textwidth]{Figures/jss2014-betapost}
+ \includegraphics[width=\textwidth]{Figures/jss2014-betapost-1}
  \caption{Visualization of the posterior distributions $\betab|\yb$ for the model with SV regression residuals, the model with GARCH(1,1) regression residuals, and the model with homoskedastic regression residuals.
@@
  \begin{center}
-  \includegraphics[width=\textwidth]{Figures/jss2014-qqplot}
+  \includegraphics[width=\textwidth]{Figures/jss2014-qqplot-1}
  \end{center}
@@
 \begin{center}
- \includegraphics[width=.984\textwidth]{Figures/jss2014-predlik1}
+ \includegraphics[width=.984\textwidth]{extrafig/predlik1}
 \end{center}
@@
 \begin{center}
- \includegraphics[width=.984\textwidth]{Figures/jss2014-predlik3}
+ \includegraphics[width=.984\textwidth]{extrafig/predlik3}
 \end{center}

### 2021-05-07 article2.Rnw
both vignettes compile and are transferred to the package during building

@@
 
-\graphicspath{{Figures/}}
+\graphicspath{{Figures2/}}
 
@@
                       cache = FALSE,
-                      fig.path = "Figures/jss3813-",
+                      fig.path = "Figures2/jss3813-",
                       tidy = FALSE)
@@
 
-\bibliography{ref}
+\bibliography{ref2}
 

### 2021-05-11 article.Rnw
decrease vignette size

@@
 
+\DeclareGraphicsExtensions{%
+  .png,.PNG,%
+  .pdf,.PDF,%
+  .jpg,.mps,.jpeg,.jbig2,.jb2,.JPG,.JPEG,.JBIG2,.JB2}
+
 \input{macros.tex}
-
-\graphicspath{{extrafig/}}
 
@@
                       error = FALSE,
+                      dev = c("png", "pdf"),
+                      dpi = 90,
                       message = FALSE,
                       echo = FALSE,
-                      cache = FALSE,
+                      cache = TRUE,
                       fig.path = "Figures/jss2014-",

### 2021-05-11 article2.Rnw
decrease vignette size

@@
 \usepackage{xparse}  % smarter commands
+
+\DeclareGraphicsExtensions{%
+  .png,.PNG,%
+  .pdf,.PDF,%
+  .jpg,.mps,.jpeg,.jbig2,.jb2,.JPG,.JPEG,.JBIG2,.JB2}
 
@@
                       error = FALSE,
+                      dev = c("pdf", "png"),
+                      dpi = 90,
                       message = FALSE,
                       echo = FALSE,
-                      cache = FALSE,
+                      cache = TRUE,
                       fig.path = "Figures2/jss3813-",
@@
   \centering
-  \includegraphics[width=\textwidth]{jss3813-svlplot-1.pdf}
+  \includegraphics[width=\textwidth]{jss3813-svlplot-1}
   \caption{The default plot of an estimated model.

### 2021-07-12 src/utils_latent_states.h
update year

@@ -1,21 +1,21 @@
 /*
  * R package stochvol by
- *     Gregor Kastner Copyright (C) 2013-2020
- *     Darjus Hosszejni Copyright (C) 2019-2020
- *  
+ *     Gregor Kastner Copyright (C) 2013-2021
+ *     Darjus Hosszejni Copyright (C) 2019-2021
+ *
  *  This file is part of the R package stochvol: Efficient Bayesian
  *  Inference for Stochastic Volatility Models.
- *  
+ *
  *  The R package stochvol is free software: you can redistribute it
  *  and/or modify it under the terms of the GNU General Public License
  *  as published by the Free Software Foundation, either version 2 or
  *  any later version of the License.
- *  
+ *
  *  The R package stochvol is distributed in the hope that it will be
  *  useful, but WITHOUT ANY WARRANTY; without even the implied warranty
  *  of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the GNU
  *  General Public License for more details.
- *  
+ *
  *  You should have received a copy of the GNU General Public License
  *  along with the R package stochvol. If that is not the case, please
  *  refer to <http://www.gnu.org/licenses/>.
@@ -23,10 +23,10 @@
 
 /*
  * utils_latent_states.h
- * 
+ *
  * Constants and utility functions related to the sampling of the
  * latent states or the mixture indicators.
- * 
+ *
  * The functions are separated into the fast_sv and the general_sv
  * variants due to the varying expert options.
  */

### 2021-07-12 tests/testthat/test-cpp.R
ready for CRAN submission

@@ -1,4 +1,3 @@
-context("C++")
 test_that("Catch unit tests pass", {
     skip_on_cran()
     expect_cpp_tests_pass("stochvol")

### 2021-07-12 tests/testthat/test-samplers.R

@@ -47,6 +47,9 @@ test_that("vanilla SV passes Geweke test", {
 })
 
 test_that("general SV passes Geweke test", {
+  skip_on_cran()
+  skip_on_ci()
+
   set.seed(60)
   result <- .Call(`_stochvol_geweke_general_cpp`, PACKAGE = "stochvol")
   draws <- result$draws / result$thin
@@ -106,6 +109,9 @@ test_that("general SV passes Geweke test", {
 })
 
 test_that("default fast SV is efficient", {
+  skip_on_cran()
+  skip_on_ci()
+
   set.seed(61)
   n <- 150L
   # Pro-centered
@@ -147,6 +153,9 @@ test_that("default fast SV is efficient", {
 })
 
 test_that("default general SV is efficient", {
+  skip_on_cran()
+  skip_on_ci()
+
   set.seed(61)
   n <- 150L
   # Pro-centered
